Effects of growth feedback on gene circuits: A dynamical understanding

Version 1

Abstract

Abstract

The successful integration of engineered gene circuits into host cells remains a significant challenge in synthetic biology due to circuit-host interactions, such as growth feedback, where the circuit influences cell growth and vice versa. Understanding the dynamics of circuit failures and identifying topologies resilient to growth feedback are crucial for both fundamental and applied research. Utilizing transcriptional regulation circuits with adaptation as a paradigm, we systematically study 435 distinct topological structures and uncover six categories of failures. Three dynamical mechanisms of circuit failures are identified: continuous deformation of the response curve, strengthened or induced oscillations, and sudden switching to coexisting attractors. Our extensive computations also uncover a scaling law between a circuit robustness measure and the strength of growth feedback. Despite the negative effects of growth feedback on the majority of circuit topologies, we identify a few circuits that maintain optimal performance as designed, a feature important for applications.

Introduction

In biomedical science and engineering, artificially designed gene circuits are anticipated to play an ever-increasing role in disease diagnosis and therapy (Riglar and Silver, 2018; Sedighi et al., 2019; Xia et al., 2019). Gene circuits also show great potential in various applications such as microbiome modulation (Foo et al., 2017; Lee et al., 2018) and biological containment (Gomaa et al., 2014; Caliando and Voigt, 2015). While most gene circuits are designed to function after they are inserted or embedded into host cells, the interactions between the circuit and the host environment are generally extremely complex and can lead to undesired effects that were not present in the original, isolated circuit (Tan et al., 2009; Ceroni et al., 2015; Borkowski et al., 2016; Ceroni et al., 2018; Darlington et al., 2018a,b; Gouda et al., 2019; Zhang et al., 2021, 2020; Melendez-Alvarez et al., 2021). Understanding the interactions and identifying the circuit topological structures that can withstand the interactions and thrive in the host are thus of fundamental importance, requiring interdisciplinary efforts among systems and synthetic biology, metabolic engineering, nonlinear dynamics and complex systems.

Typical circuit-host interactions include metabolic burden, cell growth, and resource relocation or competition, among which growth feedback is the most common type of circuit-host interaction between the circuit gene expressions and cell growth. More specifically, a synthetic gene circuit embedded in a host cell possesses an intrinsic coupling mechanism: the circuit affects cell growth and the growth in turn modifies the gene expressions in the circuit (Klumpp et al., 2009; Klumpp and Hwa, 2014; Boo et al., 2019; Scott et al., 2010) - the so-called growth feedback. Studies have shown that the growth-mediated feedback can endow a synthetic gene circuit with various emergent properties, such as innate growth bistability (Deris et al., 2013). For example, a non-cooperative positive autoregulation system, when coupled with growth feedback, gains increased effective cooperativity, thereby resulting in bistability (Tan et al., 2009; Nevozhay et al., 2012). In another example, toxin cooperativity can be induced in multiple toxin-antitoxin systems by growth-mediated feedback (Feng et al., 2014). The number of steady states in one gene circuit also depends on growth feedback and resource availability (McBride and Del Vecchio, 2020; Melendez-Alvarez and Tian, 2022). In general, growth feedback acts to hamper the forward engineering of the circuit functions by introducing modes of nonmodularity and reducing the predictability of the circuit components in an in-vivo context. While various phenomena caused by growth feedback were studied with desirable or undesirable effects on the functions of the gene circuits, a systemic picture is lacking on what effects growth feedback can have on the gene circuits, including failures.

A recent study has revealed that growth feedback can have drastically different effects on congruent circuits with distinct topologies (Zhang et al., 2020). In particular, the dynamical behaviors of two bistable synthetic memory circuits were studied: a self-activation switch incorporating positive autoregulation and a toggle switch incorporating double-negative regulatory motifs. It was found that growth feedback impacts both circuits but with quite different manifestations. For the toggle switch, memory can be retained and the circuit tends to be refractory towards growth feedback. However, for the self-activation switch, growth feedback leads to memory loss. While these results indicate that the circuit topology can play a significant role in the circuit functions when growth feedback is present, they were obtained through two specific circuit topologies. Since a particular function of the gene circuit can often be achieved by a finite set of core topologies, it is of fundamental interest to identify the most robust topologies in response to growth feedback. The so-identified optimal topologies can then be used to construct synthetic gene circuits capable of maintaining the essential functions to meet the design goals under the fluctuating growth conditions of the host cell. A systematic study of the interplay between the gene circuit topology and growth feedback is needed.

Adaptation is an important and widely studied functionality of gene circuits, which is defined as the ability of the system to respond to environmental changes and to return to the basal or near-basal state after some time (Knox et al., 1986; Tyson et al., 2003; Friedlander and Brenner, 2009; Ferrell Jr, 2016). Previously, it was found that certain circuits possess biochemical adaptation (Ma et al., 2009) if they contain at least one of the two architectural classes: an incoherent feed-forward loop with a proportion node and a negative feedback loop with a buffering node. A number of synthetic gene circuits were proposed or constructed to achieve adaptation (Kim et al., 2014; Briat et al., 2016; Aoki et al., 2019). Quite recently, a design principle for circuits with four genes was uncovered for simultaneously achieving noise attenuation and adaptation: the circuit must have a sequential assembly structure (Qiao et al., 2019). However, these existing adaptation studies did not include any growth feedback mechanism.

In this paper, we conduct a comprehensive computational study to uncover and understand the effects of growth feedback on the gene circuits. Specifically, we focus on a type of transcriptional regulation circuit designed for adaptation. There are 425 possible circuit structures (identified by previous research (Shi et al., 2017)), and we study all of them to simulate and test their response under different levels of growth feedback. Altogether, 2 × 105 sets of circuit parameters are randomly sampled for each structure. Our results reveal a vast number of cases where growth feedback has a detrimental effect on circuit function (1.3 × 105 cases in total) with varying response curves and dynamical behaviors. To gain a more intuitive overall picture, we classify these cases into several distinct categories based on the circuits’ dynamic behavior. We then systemically summarize the dynamical mechanism behind these growth-induced circuit malfunctions. To quantify circuit adaptation in the presence of growth feedback, we propose a robustness measure that enables us to identify an optimal group of circuits that exhibit a high level of robustness against growth feedback, making them particularly promising for real-world implementation. The motifs associated with this optimal group are found through machine learning. We also obtain a scaling law governing the dependence of this measure on the level of growth feedback and provide a mathematical analysis to gain insights into the underpinnings of the scaling law. The take-home message is that, in spite of the negative effects of growth feedback in the majority of the circuits, there exists a small set of circuits that are still able to deliver optimal performance as designed, which is promising for real-world implementation.

Results

A systemic search of functional failures due to growth feedback

Adaptation is referred to as the ability of a gene circuit to respond to changes in input and then to return to the pre-stimulus output level, even when the input change persists (Ma et al., 2009). More precisely, with an input signal switched from a lower value I1 to a higher value I2, as demonstrated in Fig. 1(b), a circuit with functional adaptation should have the following response-curve criteria: (i) precision - the final state O2 should be close to the initial state O1, (ii) sensitivity - there should be a relatively high Opeak in response to the change in the input, and (iii) the system should reach equilibrium within a reasonable relaxation time. A three-node gene circuit can achieve adaptation (Ma et al., 2009), with one node receiving the input (node A), another node realizing various regulatory roles (node B), and a third node outputting the response (node C). A representative circuit topology is shown inside the red dashed box in Fig. 1(a). We restrict our study of the class of transcriptional regulatory networks (TRNs) with the AND logic.

Schematic illustration of a synthetic gene circuit embedded in a host cell. (a) A representative three-gene circuit (inside the dashed red box) and its dynamical interplay with host-cell growth. Arrows with triangular ends and round ends denote activating and inhibiting regulations, respectively. Altogether, there are 16,038 possible three-node topologies, with 425 topologies capable of adaptation.(b) An example of the circuit input and output signals. The input is an idealized step function of currents I1 and I2 before and after the jump, respectively. The output signal is a response of the circuit to the step function. The features of the output signal, as characterized by three key quantities characterizing the signal: O1, O2, and Opeak, can be used to determine if the circuit has succeeded or failed in its intended function.

In our work, we use a parameter kg to control the strength of growth feedback, which is a parameter determining the maximal growth rate of the host cells, as mathematically explained in Model description. With all the other parameters fixed, a larger kg implies a faster cell growth rate and a stronger impact of growth feedback. Previous research identified 425 different three-node TRN network topologies that can achieve adaptation in the absence of growth feedback (Shi et al., 2017), providing the base of our computational study. These topologies can be classified into two families based on the mechanism they rely on to achieve adaptation: networks with a negative feedback loop (NFBL) and networks with an incoherent feed-forward loop (IFFL) (Shi et al., 2017). To investigate the effect of growth feedback on these circuits, we systematically simulate the response of the 425 network topologies under a switch in the input signal. A three-node gene circuit subject to growth feedback has a large number of parameters, which determine the properties of the regulation links within the circuit and the circuit dynamics. For each topology, we randomly sample 2 × 105 trials of circuit parameters. Altogether, our study involves analyzing approximately 8.5 × 107 different circuits. We find that among these trials, only about 1.5 × 105 meet the adaptation criterion in the absence of growth feedback. For these functional trials, we vary the growth feedback parameter kg with a series of values, and find that the majority of trials (1.3 × 105 trials, about 87%) lose their adaptation in the interval of kg ∈ (0, 1), while only 13% of trials remain functional at kg = 1.0.

A systemic classification of functional failures due to growth feedback

An essential step towards understanding the detrimental or even destructive effects of growth feedback on circuit functioning is to identify the distinct failure scenarios. Our extensive simulations have yielded a comprehensive picture of these scenarios, as shown in Fig. 2. Overall, we have identified six failure scenarios that encompass more than 99.6% of the 1.3 × 105 failing cases we collect. The first level of classification distinguishes between failures that occur continuously or abruptly as the growth-feedback strength kg increases. In a continuous failure, the response curve deforms continuously as kg increases, as exemplified in Figs. 2(a-c). In an abrupt failure, the response curve exhibits a sudden change as kg increases through a critical value, as illustrated in Figs. 2(d-f). At the next classification level, we further divide the failures into three types of continuous failures and three types of abrupt failures.

Systemic classification of circuit failure scenarios due to growth feedback. This study identifies six computationally detectable categories of failures based on the criterion of functional adaptation that the circuit violates as the effect of growth feedback becomes stronger. (a) Type-I and (b) type-II failures correspond to the cases where the precision criterion or sensitivity criterion is violated in a continuous fashion as the growth-feedback strength kg increases, respectively. (c) Type-III and (d) type-V failures occur when the circuits lose adaptation due to growth-feedback-induced oscillation, either continuously or abruptly, as kg increases, respectively. The abrupt changes in type-V are caused by bifurcations, mostly a saddle-node bifurcation of cycles or an infinite-period bifurcation. For instance, the case shown in (d) undergoes an infinite-period bifurcation. (e) Type-V and (f) type-VI failures are when the circuits lose adaptation due to an abrupt change in O1 or O2 as kg increases, respectively, which are caused by bistability or bifurcations in the systems. Trials that are not categorized under these six classifications or fall into multiple categories constitute less than 0.4% of all cases (see text for more details and discussions about each failure class). The insets around the pie chart provide exemplary response curves of the circuits in each failure scenario. Each inset shows the concentration of the output node C versus time with two values of the growth-feedback strength kg, one below and another above the failure threshold, for the specific failure scenario. In each case, the input is switched from state I1 to state I2 at the time indicated by the red vertical dashed line.

The three types of continuous failures, denoted as types I-III as illustrated in Figs. 2(a-c), are determined according to the specific quantitative criteria in the response curve that the circuits violate. Type-I continuous failures, as shown in Fig.2(a), are associated with the violation of the precision criterion. A circuit is deemed as precise if a change in the input signal (e.g., from I1 to I2) generates two opposite dynamical effects in the circuit that cancel each other out after a transient and return the final output to the original state, i.e. O2 ≈ O1. For example, in some networks [e.g., the network in Fig. 1(a)], an increase in the input signal I will result in an increase in the concentration of gene A and a reduction in the concentration of gene B. As both genes regulate the output gene C with the respective activation links, for proper system parameter values, the two effects will cancel each other out, resulting in O2 ≈ O1. Type-I continuous failures constitute the largest failure category among all possible circuit topologies, suggesting that the exact cancellation is fragile and the loss of precision is the most common dynamical mechanism behind growth-feedback-induced failures.

Our simulations reveal that an exact cancellation between the two opposite sources at kg = 0 prevents an exact cancellation at any other values of kg. That is, the set of circuit parameter values leading to perfect precision, in general, depends on the value of kg (see Appendix 1 for more details). The implication is that, for fixed circuit parameter values, achieving high precision under growth feedback (kg > 0) is difficult if the circuit is precise in the absence of growth feedback (kg = 0).

Type-II continuous failures are characterized by a continuous change in the peak of the response curve, denoted as Opeak, as kg increases, eventually falling below a threshold, as shown in Fig. 2(b). This type of failure can make it challenging for downstream circuits to detect the peak signal, hindering information transmission in the larger system. Type-II failures are the second most common type of failure observed in our simulations. The occurrence of a high peak in the response curve requires a significant transient deviation from the final equilibrium point. In the presence of growth feedback, the transient behavior changes, which can further alter the peak height Opeak.

Type-III and type-IV failures arise due to growth-feedback-induced oscillations, while type-V and type-VI failures are caused by bistability or bifurcations, with fold bifurcations being the most common type. To provide a more detailed understanding of these different failure scenarios, we discuss the two mechanisms, respectively, in the sections of Growth-feedback induced oscillations and Bistability and bifurcations below.

Three classes of growth-induced failures

All the failures we observed can be categorized into the following three general classes, applicable to both the three-gene and four-gene circuits we tested:

Continuous Deformation of the Response Curve Typically, we require a specific range of response curve shapes for a gene circuit, such as a peak in the output with a minimum height or duration. In a failure caused by continuous deformation, the growth feedback prompts a gradual change that crosses the boundary of these criteria for response curve shapes.

Growth-Induced or Growth-Strengthened Oscillations Growth feedback can induce oscillations in a circuit through various types of bifurcations or amplify existing oscillatory behavior with longer relaxation times or larger amplitudes. A circuit experiencing growthinduced or growth-strengthened oscillations cannot reach a relatively steady state (an equilibrium or relatively small oscillations) within a finite time or reasonable relaxation period.

Growth-Induced Switching Among Coexisting Attractors When coexisting attractors are present in the circuit dynamics, such as bistability or multistability, the circuit typically only functions with one of the attractors. In other words, the circuit is functional locally in its phase space rather than globally. Strengthened growth can push the system across the boundary of different attracting basins in the circuit phase space, causing the circuit to lose its desired functionality by switching from a functional basin to a malfunctioning basin.

Growth-feedback induced oscillations

As demonstrated by the light green and yellow slices of the pie chart in Fig. 2, a considerable portion (17%) of the circuit failures are caused by growth-feedback-induced oscillations. Growth feedback perturbations can easily change the system from the adaptive domain to the oscillation domain in these cases. Our program classifies oscillation-mediated failures into two categories: continuous (type-III) and discontinuous failures (type-IV). Type-III failures are the results of either (i) a gradual increase in the oscillation amplitude, or (ii) a gradual increase in the transient lifetime of damped oscillations. In the first case, an isolated circuit has already exhibited oscillations with small amplitudes in its gene concentrations with relatively weak growth feedback. As the feedback is strengthened with a larger value of kg, the oscillations are intensified with a larger amplitude, leading to a circuit failure. In the second case, there is damped oscillation for small kg with a relatively short transient time before approaching an equilibrium. After strengthening the growth feedback, the damping weakens and the oscillation’s amplitude cannot be reduced to the threshold within the time limit, as exemplified in Fig. 2(c).

The second category of growth-feedback-induced oscillation is type-IV, where oscillations emerge suddenly as the growth-feedback strength increases through a critical point. The sudden emergence of oscillations can be caused by a bifurcation or a transition into a basin of a limit-cycle attractor. A random sampling of the type-IV failure cases reveals that most of them are caused by either a saddle-node bifurcation of cycles (Strogatz, 2018) or an infinite-period bifurcation (Stro- gatz, 2018). In the former case, a pair of stable and unstable limit cycles suddenly emerge together. In the latter case, when observed from the opposite direction (i.e., with a decreasing kg crossing the threshold), the oscillation in the system spends a longer and longer time around a node on the limit cycle. This node finally becomes a stable fixed point at the bifurcation point, and the oscillation period approaches infinity. One example of type-IV oscillation-mediated failures caused by an infinite-period bifurcation is shown in Fig. 2(d). In our simulations, most of the cases where there are saddle-node bifurcations of cycles are categorized as type-III failures because, prior to the bifurcation point, the system can be oscillating near the ghost (Strogatz, 2018) cycle for a long time exceeding the criterion for relaxation time, though that ghost cycle is not an attractor but only a transient in the system.

Our results indicate that for various circuit topologies, the dynamic mechanisms leading to failures can differ, resulting in significantly different distributions of failure types among different networks. For instance, the fractions of failures caused by growth-induced oscillations can vary dramatically among all the topologies, as demonstrated in Fig. 3(a), where each data point represents a specific network topology. The fraction of failures caused by growth-induced oscillations can range from approximately zero to about 80%! A particular example of two different networks is presented in Figs. 3(b1) and 3(b2), both of which share the same minimal topology required for adaptation (Shi et al., 2017) - the circuit’s core function. Despite differing by only one link, the proportions of failures with unique mechanisms are quite distinct, as illustrated in Figs. 3(c1) and 3(c2). Notably, for the network in Fig. 3(b1), almost half of the failures result from oscillations, while hardly any oscillation-mediated failures occur for the network in Fig. 3(b2). The explanation is that, although the difference lies in only a single link, this link determines whether an oscillation-correlated motif exists within the network. Previously, three classes of motifs capable of supporting persistent oscillations were discussed (Novák and Tyson, 2008), including the “delayed negative-feedback loop” featured in Fig. 3(b1).

Frequencies of growth-feedback induced oscillation failures for different network topologies. (a) Significant variations in the proportion of trials resulting in circuit failures due to growth-feedback-induced oscillations (types III and IV failures) across distinct network topologies. Some topologies exhibit virtually no oscillation-related malfunctions, while others experience about 80% of failures caused by growth-induced oscillations. Network topologies containing any oscillation-supporting motifs (discussed in the main text) are represented by red triangles, while the rest are shown as blue circles. The majority of red data points have higher fractions of oscillation-related failures compared to the blue ones, mainly due to the presence of oscillation-supporting motifs. To reduce fluctuations in the results, only circuit topologies with over 200 failed trials are included. (b1, b2) A pair of network topologies that differ by only one link (from node C to node B). (c1, c2) The distinct topologies in (b1, b2) leading to different distributions of failure mechanisms. The topology in (b1) primarily experiences growth-induced oscillation as the major failure mechanism, while the one in (b2) has barely any trials with growth-feedback induced oscillations.

Generally, the circuit dynamics depend sensitively on the structure, but oscillations specifically require a negative feedback loop with time delay (Novák and Tyson, 2008). Since there are no explicit time-delayed terms in the dynamical equations in our model, one of the two types of motifs - an intermediate node in the path of the negative feedback loop or an additional positive feedback loop - is necessary to induce time delay (Novák and Tyson, 2008). For network topologies with a high ratio of functional failures caused by oscillations, both motifs are observed, especially the former type. For the network in Fig. 3(b1), the three links: A → C, C ⊣ B, and B → A, together constitute a negative feedback loop, making the circuit more susceptible to oscillatory behaviors. For the circuit in Fig. 3(b2), no such negative feedback loop exists. Figure 3(a) summarizes the total number of failed trials and the ratio of oscillation-induced failures for each network topology. The network topologies that contain one of the motifs for oscillation as discussed in Novák and Tyson (2008) are marked in red, while the networks that do not consist of any of them are marked in blue. Note that all the networks with relatively high ratios of oscillation-induced failures (e.g., ratio > 0.2) consist of oscillation-correlated motifs. Details about these oscillation-correlated motifs are discussed in Appendix 2.

We conclude that, for a network with an oscillation-correlated motif, even if it is functional at some parameter values, the potential of oscillatory behaviors can be triggered by growth feedback. As a result, networks without these motifs can be safer choices to avoid too many failures cases due to oscillations. Note that this relationship is not deterministic. As shown in Fig. 3, even the networks represented by blue dots that have no oscillation-correlated motifs can still have oscillation-induced failures (with small ratios). The complexity of the scenario makes it challenging to find general and relatively simple rules that connect circuit topology to the circuit’s robustness.

Bistability and bifurcations

In this section, we describe the dynamical mechanisms behind type-V and type-VI failures, which in total take up about 14% of all the circuit failures. These failures are abrupt, meaning that the response curve undergoes an abrupt change at a certain critical value of kg from a desirable curve of adaptation. Type-V and type-VI failures correspond to an abrupt change in O1 and O2, respectively. Both types of failures are closely related to bistability and bifurcations.

Bistability and multistability are common phenomena in nonlinear systems. Bistability refers to the situations where two stable attractors coexist in the phase space simultaneously. Multistability describes a similar coexisting phenomenon of attractors, but with more than two attractors. With bistability or multistability in the target dynamical system, the system trajectory may end up in any one of these stable attractors, depending on the initial state of the system evolution. The entire phase space can thus be separated into two or more basins of attraction. Each basin of attraction corresponds to an attractor and consists of all the initial states that eventually lead the system to the attractor. The boundary boundaries separate two different basins of attraction. A close pair of initial states but at different sides of a basin boundary lead the system to different attractors.

In our simulations, we observe bistability in many circuits. While multistability has also been observed, it is relatively rare. We thus focus on bistability. It is highly unusual for both attracting basins to exhibit the desired functional behavior simultaneously. This is because they are located in different regions of the system phase space, and accommodating both would impose overly stringent constraints on the circuit. Having one basin functional is already difficult enough with a random sampling of circuit parameters. As a result, functional adaptation is typically found in only one of the basins, with adaptation being lost in the other, and the circuit is functional only locally in its phase space, rather than on a global scale. A drifting system parameter, such as kg, can alter the dynamics of the gene circuit. In a situation with bistability, such a change in the system dynamics can modify the shape and position of the basin of attraction and the basin boundary. Consider an initial state close to a basin boundary. With the deformation caused by a drifting parameter, the boundary may shift across the initial state, leading to a sudden switching of the system’s final attractor. If the basin before the parameter change is functional and the basin after is not, this leads to a growth-feedback-induced failure. The crossing of the basin boundary dictates that the system’s final state will abruptly change from one attractor to another. This type of failure can be classified as a switching type of failure.

An example of bistability-related failures is shown in Fig. 2(e), where in the upper panel, the circuit enters into the functional region after an initial transient. In the lower panel, the circuit enters into another region that does not have adaptability, and the circuit does not respond to the switching of the input signal. Figures 4(a) and 4(b) illustrate how the basin structure of the circuit changes significantly with different values of kg. The functional basin is in yellow and it shrinks greatly with an increasing kg. Note that the phase space is four-dimensional, so only a two-dimensional slice is shown. For a bistability/multistability-induced type-V failure where O1 is switched, the boundary of the functional basin crosses the initial state. For a type-VI failure, the simultaneous movement of both O1 and the basin boundary under input I2 results in O1 crossing the boundary.

Two possible mechanisms behind both the type-V and type-VI failures. (a,b) Basin structure of the circuit shown in Fig. 2(e) (with type-V failure) for input I1 with different levels of growth feedback, for growth feedback strength kg = 0.05 (weak) and kg = 0.97 (relatively strong). The coordinates A0 and B0 are the initial values of nodes A and B, respectively, corresponding to a two-dimensional slice of the entire four-dimensional phase space by fixing C = 0.1 and N = 10−3. The color bar indicates the equilibrium value of node C before the input switch, which is O1(C). There is bistability in both cases, as there are two basins of attraction. The yellow region is the functional basin that has adaptation, while the blue region is a non-functional basin without adaptation. The relative size of the blue non-functional region with larger kg in this case is significantly larger and includes the initial state of the system (A0 = B0 = 0.1), causing a type-V circuit failure. (c) Bifurcation diagram of the circuit in Fig. 2(f) with a type-VI failure. Prior to a critical value of kg ≈ 0.0122, only one stable value of O2(C) exists. There is a fold bifurcation at the critical value, giving birth to a new state around zero, which is non-functional. For an abrupt failure, whether it belongs to type-V or type-VI does not determine whether it is caused by bistability or bifurcations. Both mechanisms are possible for both types of failures.

Bifurcations also play an important role in many type-V and type-VI failures. An example of a failure caused by a fold bifurcation is shown in Fig. 2(f) and the corresponding bifurcation diagram is shown in Fig. 4(c), where a non-functional fixed point appears through a fold bifurcation as kg crosses a critical value. Bifurcation-induced abrupt failures differ from those caused by bistability/multistability, but they can be related since significant changes in the basin structures often occur near a bifurcation point.

Circuit robustness and optimal topology

To quantify a circuit’s robustness against growth feedback, we introduce two metrics: Q-value and R-value. We track the number of remaining functional trials for each network for various kg values (starting from kg = 0), denoted as Q(kg). This measure extends the concept of Q-values in Ma et al. (2009) by accommodating non-zero values of kg. To characterize the circuit robustness, we define the survival ratio R(kg) as R(kg = k) = Q(kg = k)/Q(kg = 0). This ratio represents the fraction of random circuit realizations that maintain functionality under growth feedback with a strength of kg.

Note that each Q(kg) or R(kg) is defined for a specific network topology in a suitable parameter space. A high value of R(kg) indicates that a large fraction of the randomly sampled circuit parameters is functional despite cell growth with any strength no larger than kg, indicating that the topology is more robust against growth feedback. Because of the detrimental effects of growth feedback, R(kg) decreases monotonically with respect to kg.

To justify the utility of R(kg), we test the circuit topologies employed in a previous work (Zhang et al., 2020), where two relatively simple network topologies were used for a comparison study in terms of their ability to resist growth feedback and remain functional. Our evaluation of R(kg) for the two topologies has yielded results that are consistent with those in Zhang et al. (2020), as discussed in Appendix 3. To illustrate our results in a concrete way, we set kg = 0.6 and calculate the ratio R(kg = 0.6) for different network topologies.

Our computations have revealed a set of eight circuit topologies with optimal performance as characterized by high values of both Q(kg = 0) and R(kg), as indicated by the set of orange points in Fig. 5(a). The optimal circuits form a family as their topologies exhibit a high level of similarity with one other. In particular, all eight circuits in this family share a common set of links (motif), as shown in Fig. 5(b). The combination of these common links is one of the minimal topologies with perfect adaptation in a three regulatory logic (Shi et al., 2017) and is critical for the circuit to be functionally adaptable. The only difference among the circuits in this family is the links from node C. While an inhibition link from node C can be important to achieving a value of R(kg), as discussed below, the eight optimal circuit topologies do not contain any such inhibition link from node C. (The role of this particular link will be further studied in our analysis of the results in Fig. 6.) This also explains why the family has eight members, as follows. Each link from C has two options: either the link does not appear, or it appears as an activation link. As there are three possible links from C (C to A, C to B, and C to C), there are altogether eight (23) topologies within this optimal family, according to the simulation results in Fig. 5.

A family of circuit topologies with optimal performance. The circuits both have a large volume of the functional region in the parameter space in the absence of growth feedback as characterized by a large value of Q(kg = 0) and are robust against growth feedback with a high value of R(kg). (a) Values of Q(kg = 0) and R(kg = 0.6) from all the 425 network topologies, where each data point corresponds to a topology. The family of optimal topologies is represented by the orange data points, including eight network topologies. (b) The set of links (motif) shared by this family of circuits. The combination of these links is also one of the minimal topologies with perfect adaptation in three regulatory logic (Shi et al., 2017).

Strong correlation between circuit robustness against growth feedback and circuit topology. There are three groups of circuits, each displaying strong topological similarities within, exhibit distinct levels of robustness against growth feedback as measured by the characterizing quantity R. (a) Robustness measure R(kg = 0.6) versus Q(kg = 0) for all 425 network topologies. Circuits are color/shape-coded into three groups (green triangles, blue circles, and red diamonds) based on the rules defined in the text. The three groups of topologies display distinct levels of R(kg = 0.6) values, signifying a strong correlation between circuit robustness and topology. Only circuits with Q(kg = 0) > 300 are shown to reduce fluctuations arising from random parameter sampling. What is demonstrated is the case of an intermediate level of growth feedback with kg = 0.6 (a different value of kg has no significant effect on the results - see Fig. 7). The topologies associated with the green triangles have a high level of robustness, which can be regarded as an optimal group and is more prevalent than the optimal group identified in Fig. 5. (b) Histogram of R(kg = 0.6) the same color legends as in (a). Three distinct peaks emerge, each associated with a group of circuit topologies. (c) The shared network motif among all networks in the green group, which is highly correlated with the optimal minimal network shown in Fig. 5(b), but without the link B → B, which is necessary for the NFBL family of networks to have adaptation (Shi et al., 2017). (d) Effects of burden b for the three groups of networks, where the abscissa is the effective term of burden in the formula of growth rate Eq. (12). The circuits in the red group have larger values of 1/(1 + ⟨b⟩), suggesting that a heavier burden yields a stronger effect of the growth feedback for the red group. (e) A multilayer perceptron (MLP) for identifying the crucial connections that determine the robustness of the circuits. The circuit topology serves as the input, where 1, 0, and -1 represent activation, null, and inhibition links, respectively. The output is a predicted robustness measure, denoted as . To encourage the neural network to select as few links as possible for predicting , a l-1 regularization term, β||Win||, is incorporated into the loss function alongside the fidelity error . As a result, the feed-forward process eliminates information about the links that have little impact on circuit robustness since the corresponding Win entries automatically optimize to values close to zero. (f) Results from an ensemble of 50 MLPs, each trained with distinct initial values. Shown is the average importance of each of the nine links, which is determined by the weights in Win - see Appendix 4. The top four links with the highest importance correspond to the four links used to classify the three peaks in panel (b).

How can we quickly determine if a three-gene regulatory network with a given topology can be robust against growth feedback? Is there any structural feature of the circuit that can be used to estimate if a high value of R(kg) can be achieved? To gain insights, we observe that the histogram in Fig. 6(b) has three peaks about low, moderate, and relatively high values of R, respectively. Computations reveal certain “shared topological similarity” (or motif) within each peak. Thus, each peak corresponds to a group of similar network topologies that simultaneously have a similar level of R. This observation suggests a correlation between the network topology and robustness against growth feedback. For convenience, we refer to these three groups of networks by the colors presented in Fig. 6. For instance, the group with the highest R (the green triangles in Fig. 6(a)(d)) is called the green group, and the group with the lowest R (the red diamonds in Fig. 6(a)(d)) is the red group.

To better distinguish the three groups, we introduce two binary variables, B1 and B2. For each network, B1 = 1 if the network contains the motif in Fig. 6(c), and B1 = 0 otherwise. Then, for each network, an additional binary variable is set to be B2 = 1 if there is an inhibition link from the output node C, and B2 = 0 otherwise. We find that a linear combination of the two binary variables, BS = B1 − B2, can characterize the circuit topology and robustness against growth feedback. In particular, the three possible cases BS = 0, 1, or -1 correspond to the three peaks in Fig. 6(b). This result suggests that the motif shown in Fig. 6(c) is beneficial for robustness, while an inhibition link from the output node C is detrimental. It is the balancing act of these two factors that determines the overall circuit robustness.

The discovery of this three-peak structure and the corresponding topological similarity within each peak is facilitated with the use of machine learning. In particular, we consider a simple type of artificial neural network called multilayer perceptron (MLP), where we train it to predict the R value from the input of the network topology through a small hidden layer with only two nodes, as demonstrated in Fig. 6(e). This bottleneck structure in the hidden layer plus the l-1 regularization imposed on the input matrix Win forces the MLP to extract low-dimensional features from the input topology to estimate R. In our tests, the MLP designed this way automatically assigns different levels of weights to the input information of different links. Over an ensemble of 50 MLPs trained with different random initial values, the ranking of average importance is shown in Fig. 6 (f). The top four links are the four links used to categorize the three peaks.

The results in Fig. 6 is for kg = 0.6. However, we find that different values of kg lead to essentially the same ranking of R(kg) among the circuit topologies, as illustrated in Fig. 7.

Robustness of the circuit division into three groups subject to different levels of growth feedback. From the top to the bottom, the four rows are for kg = 0.2, kg = 0.4, kg = 0.6, and kg = 0.8, respectively. The legends are the same as in Figs. 6(a) and 6(b). For different levels of growth feedback, the distribution of the robustness measure exhibits three distinct peaks that occur at approximately the same locations on the R axis. The implication is that the division of the circuit topologies into three groups in terms of the robustness measure can be revealed by examining the circuit functions at a single value of the growth feedback strength.

Three remarks on our categorizing rules based on the two extracted featured motifs are in order.

First, the shared motif for the green group is strikingly similar to the optimal minimal network in Fig.5(b) (the orange group). The sole distinction lies in the self-activation link of node B. This specific link plays a crucial role. Every network in the NFBL family depends on this link to achieve adaptation (Shi et al., 2017). However, for circuits within the IFFL family, this link is not a necessity for adaptivity. Missing this link makes the motif in Fig. 6(c) no longer a minimal network for adaptation, and a circuit containing this motif may either belong to the NFBL or the IFFL family. We have thus identified two optimal groups: the green group with optimal robustness R and the orange group with both the optimal robustness R and the largest functional volume Q(kg = 0) in the absence of growth feedback. The orange group is a subset of the green group, with an additional requirement for Q(kg = 0).

Second, the shared motif for the red group is also exactly the group of all circuits containing an inhibition link from node C to node B, denoted as C ⊣ B. These two different definitions are in fact equivalent: all networks with Q(kg = 0) > 300 that contain C ⊣ A or C ⊣ C also contain the motif in Fig. 6(c), yielding BS = 0, and belong to the blue group.

Third, the three circuit groups in Fig. 6 are not correlated with the categories used in previous research on circuit functionalities without growth feedback (Ma et al., 2009; Shi et al., 2017). These studies classified adaptive networks into NFBL and IFFL families. Each family contains a few minimal topologies with or without some additional other motifs, and the two families have distinct minimal functional topologies. The minimal topology acts as the backbone for supporting circuit functionality. We find that, when growth feedback is present, the prior classification scheme and the underlying minimal topologies become less relevant. Circuits belonging to the NFBL family are spread across all three levels of R(kg) in Fig. 6(b), as are the circuits from the IFFL family. A robust circuit can be part of either family, just as a fragile circuit can belong to both. We give that: (i) the topological motifs determining circuit functionality robustness and (ii) the motifs deciding whether a circuit belongs to the NFBL or IFFL family are independent. To quantify this irrelevance, we calculate the point biserial correlation between R(kg = 0.6) and a binary variable determining the family to which the circuit belongs. The resulting correlation is merely 0.1, suggesting hardly any correlations. A further illustration and quantification of this irrelevance can be found in Appendix 5.

What are the reason and mechanism behind the phenomenological set of circuit categories? Especially, it is desired to understand why the shared motif for the green group is beneficial for circuit robustness, and why the shared motif for the red group is harmful for robustness. It is challenging to find straightforward explanations given the complexity of the problem (see Discussion section). Certain insights are as follows. We find that the average node concentrations at the equilibrium for the network topologies in the red group are consistently smaller than those in the blue group. This difference is reflected in the value of burdens b. In particular, according to Eq. (12), the cell growth rate is proportional to the term 1/(1 +⟨b⟩) under the same level of growth feedback. Figure 6(d) shows the average burden b for each network topology, demonstrating that the values of the term 1/(1 + ⟨b⟩) for the circuits in the red group are larger than the values in the blue group. As a result, for the same value of kg, the growth feedback effectively received by the circuits in the red group is stronger than that of the blue group circuits. Further support is provided by the results from the limit J → ∞ (Appendix 6). In this limit, the burden b does not affect the strength of the growth feedback. As a result, the R values of the red group significantly overlap with those of the blue group, suggesting that the distinctively low values of R for the red group be a result of the burden with finite J. We also find that the existence of the shared motif for the red group has a stronger correlation to the motif necessary for growth-feedback induced oscillations. All circuits with oscillation type of failures taking up more than 20% of failures belong to the red group. This correlation can result in further fragility of the red group circuits.

Scaling law quantifying the effect of growth feedback on gene circuits

A comprehensive way to understand the effects of growth feedback on gene circuits is through scaling laws, an approach commonly employed in statistical and nonlinear physics. Does a scaling law exist that characterizes quantitatively how growth feedback affects the circuit functioning? Through a systematic computational analysis of the circuit robustness, we have uncovered a scaling law that governs how the robustness measure R(kg) deteriorates as growth feedback is strengthened, as shown in Fig. 8, where the blue curve is the result averaging over all the 425 network topologies. The three other curves represent circuits that have a relatively high, moderate, and low value R among the 425 topologies tested. As growth feedback is strengthened, the number of circuit topologies that can maintain functioning decreases (or, equivalently, the number of failed circuits increases). The decreasing behavior of R(kg) with kg tends to be slower than exponential [e.g., exp(−βkg) with β > 0 being a constant].

Scaling law governing the circuit robustness measure R(kg). The blue curve is the average result of all the 425 network topologies. The other three curves are from three types of circuits with relatively high, moderate, and low values of R among the 425 circuit topologies tested.

A general theoretical argument for the scaling law is unavailable. However, if we simplify the system by setting the parameter J in Eq. (12) to be large so the burden b(t) is much smaller than one, we are able to argue that the scaling law is approximately given by where β > 0 and 0 < λ < 1 are two specific constants that depend on the network topology, and the typical value of λ is about 0.6. The exponential scaling is assumed, given its memorylessness. That is, there is no special zero point of kg, for the reason that a certain level of kg is mathematically equivalent to a larger dx, as discussed below.

The quantity R(kg) is a simple and straight-forward measure characterizing the detrimental effect of growth feedback on gene circuits. We carry out a semi-quantitative analysis of this quantity and the effect of kg on it. The circuit dynamical equations Eqs. (9-12) can be simplified by substituting Eq. (14) into them to cancel the dN/dt terms, leading to Compared with the equations without growth feedback Eqs. (6-8), we see that introducing growth feedback is equivalent to adding a variable kg/(1 + b) to the degradation terms for each node. Intuitively, the value Q of a network topology measures the volume of the functional region ℳ in the parameter space, which is also a function of kg. We thus have that R(kg = k) is the volume of the intersection between ℳ(kg = k) and ℳ(kg = 0) divided by the volume of ℳ(kg = 0): where V (ℳ) is the volume of ℳ.

The picture can be further simplified if we assume the burden b is approximately a constant within a range of kg. Since growth feedback contributes to an additional term in degradation dx, strengthening the feedback is equivalent to increasing all three quantities dx together. Consequently, as kg increases, the high dimensional region M does not deform, but simply translates in the negative direction in all dimensions of degradation dx in the parameter space. That is, as growth feedback becomes stronger, it becomes more difficult for the circuit to maintain it functioning.

Discussion

When a synthetic gene circuit is introduced into a host cell, an inherent coupling arises wherein the gene circuit affects cell growth and cell growth in turn alters the circuit gene expression (Klumpp et al., 2009; Klumpp and Hwa, 2014). Due to the fundamental nonlinearity in the gene network and in the cell growth dynamics, the interaction is generally quite complicated. To understand this interaction so as to identify the circuit topologies that can withstand the interaction and maintaining the intended circuit functions is one of the most challenging problems in synthetic biology.

Previous studies showed that growth-mediated feedback can endow synthetic gene circuits with various emergent properties. In general, growth feedback tends to negatively impact the intended function the circuit is designed for. There was preliminary evidence that the effects of growth feedback depend strongly on the circuit topology (Zhang et al., 2020). For a particular circuit function, while the vast majority of the topologies would fall under growth feedback, a handful still exists that is adaptable to maintain its designed functions. Identifying the “optimal” topologies that are most robust against growth feedback is fundamental to constructing synthetic gene circuits that can survive, adapt, and function as designed in the fluctuating growth environment of the host cells.

The main contribution of this paper is a systematic computational study of three-gene circuits with adaptation to uncover and understand the detrimental effects of growth feedback on gene circuits and to identify optimal groups of topologies. Without growth feedback, there are 425 possible topologies with functional adaptation. A vast majority of these circuit topologies fail in their functions under growth feedback, and our computations have revealed, for the first time, six distinct main failure categories covering more than 99% of the cases. From a dynamical point of view, there are three mechanisms by which growth feedback can deprive the circuit of its ability to adapt: (i) continuous deformation of the response curve, (ii) strengthened or induced oscillations, and (iii) sudden switching to coexisting attractors (also summarized in Box 1). By introducing a robustness measure to quantify circuit adaptation in the presence of growth feedback, we uncover a general scaling law characterizing the detrimental effect of growth feedback on the circuit functioning in a quantitative manner. We identify an optimal group of circuits with high robustness and key subsets of links associated with this group that play a critical role in sustaining circuit function in host cells. Taken together, to design a functional gene circuit, growth feedback must be taken into account, as the same circuit designed with perfect functions without the feedback can behave quite differently when the feedback is present. Our study has provided unprecedentedly quantitative insights into the interplay between gene circuit topology and growth feedback, unlocking the dynamical mechanism of growth-induced failures and providing guidance to better design practically applicable synthetic gene circuits.

A unique finding is that growth feedback can induce or strengthen oscillations in gene circuits designed for adaptation. Such oscillations can often destroy the circuit functionality. In a recent experimental study, a similar phenomenon was observed in gene circuits designed for self-activation (Melendez-Alvarez et al., 2021). These results suggest that growth-feedback induced oscillation may be a general dynamical mechanism that can negatively affect the robustness of gene circuits. In addition, our study has shown that growth feedback has a highly sensitive dependence on the circuit topology: even a small structural differences between two circuits designed for the same function can result in drastically different outcomes under growth feedback. For example, Fig. 3 demonstrates that a link critical for an oscillation-supporting motif can significantly affect the robustness of the circuit against growth feedback. It can thus be quite useful to identify failure-related motifs so that they can be avoided when designing a gene circuit.

From a broad point of view, our study has yielded basic insights into the fundamental topology-function relationships in gene circuits. Examples include how circuit topology affects circuit robust-ness against growth feedback and whether a circuit topology contains motifs supporting a specific type of growth-induced failure, such as oscillation-related malfunctions. However, searching for and understanding the interplay between circuit topology and dynamical behaviors remain to be a challenge, for the following five reasons.

First, the two relevant questions are whether a circuit topology supports adaptation and whether the circuit is robust against growth feedback or is susceptible to a specific type of growth-induced failures. While our study focused on the latter, the former is important. Addressing both questions to identify and analyze all possible scenarios is infeasible at the present, due to the complex parameter space of the circuits. To make our study feasible, we focused on the cases where the circuit satisfies all the requirements for adaptation in the absence of growth feedback. These cases may occupy a small region in the entire parameter space of the circuit. For each circuit topology, the uncovered function failures due to growth feedback are thus limited to relatively small parameter regions. Second, most network topologies studied have dense connections among the three nodes (only about 20% of the networks have fewer than six connections). As a result, different motifs can overlap with each other, blocking or enhancing the function of each other. The dense connections thus pose a difficulty in identifying the motifs accurately. Third, for a particular class of failures, competition among different failure types may arise. For instance, a circuit with oscillation-supporting motifs may not have a high fraction of oscillation-induced failures because it also contains the motif for bistability, leading to a large fraction of failures due to the bistability-induced malfunctions. Fourth, due to the necessity to set a threshold in the relaxation time, transient behaviors can arise. In many failure cases caused by oscillations, the oscillatory behavior is not stable and the circuit will eventually approach a fixed point. However, time scales should be taken into account. The transient behaviors can make the network topologies without the necessary motif for sustained oscillations exhibit oscillation-induced failures. Fifth, growth feedback acts as additional feedback loops within the circuit, potentially complicating the circuit dynamics and adding more links to the circuit topology. These extra links in the integrated topology might give rise to an oscillation-related motif. However, our simulations have shown that the impact of this additional oscillation motif, introduced by growth feedback, tends to be weak (Appendix 6).

In Appendix 7, we extend our analysis to four-gene circuits with over two thousand functional failure trials. A remarkable finding is that the failure scenarios for these four-gene circuits are the same as the categories for three-gene circuits (summarized in Box 1), indicating that the growth-feedback induced failure mechanisms identified in our work are general.

The primary goal of this paper is to explore the ways in which growth feedback can undermine a gene circuit’s functionality, and we have uncovered three main dynamical failure mechanisms. The reason that we focus on small gene circuits (i.e., those with three or four genes) is that, in current synthetic biology, only small gene circuits are of interest. The main reason is that, even for a modest number of genes, when the circuit is introduced to a host, the competition and interactions in the form of growth feedback are likely to lead to unintended and uncontrollable consequences. Another reason is resource competition: the genes in the circuit could compete for the limited resources in the host cell, negatively impacting the circuit dynamics. Because of the two reasons, at present large gene circuits are not favored in synthetic biology. In fact, the state-of-the-art synthetic gene circuitry usually involves three or four genes, where the consequences of growth feedback had been poorly understood. Our work fills in this knowledge gap.

It is possible that, in the future, synthetic biology may use larger and more complex circuits. To uncover and understand the failure mechanisms as well as to identify circuits that are resilient to growth feedback, machine learning can be used. For example, recurrent neural networks have recently been used to identify circuit topologies appropriate for a specified desired function (Shen et al., 2021), and reinforcement learning tackle the combinatorial optimization problem (Bello et al., 2016; Mazyavkina et al., 2021) of pinpointing the optimal circuit topologies. Furthermore, automated differentiation (Hiscock, 2019; Kong, 2022) can be exploited to locate optimal network parameters, which can be efficient for larger circuits with a high-dimensional parameter space. In spite of these works, to study the effects of growth feedback and resource competition among numerous genes in larger circuits remains to be a formidable challenge. Our work providing a comprehensive picture of the failure mechanisms induced by growth feedback represents a step forward in this field.

Models and Methods

Model description

We restrict our study of the class of transcriptional regulatory networks (TRNs) with the AND logic. For an isolated circuit (in the absence of any growth feedback) with the topology specified inside the red dashed box in Fig. 1(a), the dynamical equations are where the dynamical variables A, B, and C are the concentrations of each protein (node). The notations are as follows. Let x and y be two arbitrary nodes. The quantity νx is the maximal production rate of gene x, dx is the degradation rate of gene x, dx/dt is its time derivative of the concentration, nxy and Kxy are the coefficients in the Hill function for a transcriptional regulation from gene x to gene y.

When growth-mediated feedback is present, the dynamical equations of the three-node circuits are modified to where the additional dynamical variable N denotes the density of the host cells, kg is a parameter controlling the maximal growth rate of the host cells, J is a parameter reflecting how this three-node gene circuit contributes to the burden.

The growth of N is under the regulatory action of two sources: by itself following the logistic equation with the environmental capacity N0 and by the burden b that represents the competence from the metabolism of the gene circuit. To make the computations feasible, we focus our analysis on the exponential growth phase so that N0 ≫ N. The equation governing the growth of the cell numbers, Eq. (12), can then be rewritten as where the dilution rate dN/dt is regulated only by the burden b(t) of the gene circuit. While cell growth is inhibited by the metabolism of the gene circuit, the circuit is also regulated by the growth of N that dilutes the concentration of circuit nodes with increasing cell volume. This dilution is reflected by the additional terms −(x/N)(dN/dt) in Eqs. (9-11).

It is useful to clarify the meaning of the degradation parameter dx and its relationship to growth feedback. While degradation and growth feedback terms have the same sign in the regulatory equations, dx may include a constant dilution. We assume that dx represents the sum of all the degradation effects in cells that are distinct from growth feedback. For instance, degradation tags, especially in the ssrA tagging systems (Gottesman et al., 1998), are often used in synthetic gene circuits to increase the degradation rate and thus increase the time scale of the whole system (Elowitz and Leibler, 2000; Fung et al., 2005; Stricker et al., 2008; O’Brien et al., 2012).

Numerical criteria for functional adaptation

We introduce four criteria to determine if a circuit has functional adaptation.

Precision

The basic requirement of adaptation is that the output remains the same when is input is switched from one state to another, i.e., O2 should be close to O1 in Fig. 19b). Specifically, we set the precision criterion to be |(O2 − O1)/O1 |< 0.1.

Sensitivity

The circuit is also required to respond to the switch of the input signal with a high peak. This ability of the circuit is named sensitivity. We introduce two types of sensitivity: relative and absolute, with the respective criteria Opeak/O1 > 0.5 and Opeak > 0.1. Only the circuits meeting both criteria are regarded as having achieved the required sensitivity.

The need to use the two different criteria simultaneously can be justified, as follows. Given the variety of network topologies and a large number of system parameters, there is a vast diversity in the circuit dynamics and the values of O1. When O1 is small, it is difficult to observe a peak that has even satisfied the relative sensitivity criterion. If the absolute criterion is used alone for a circuit with a large O1 value, the peak may be negligible in comparison with O1, making its observation practically difficult. It is thus necessary to combine the two criteria so that the cases of small and large values of O1 can be dealt with on the same footing.

Oscillations

To achieve the desired adaptation, the circuit’s output should reach a steady state before and after the input signal is switched. The values of O1 and O2 can be determined as the output signal associated with the steady states. However, realistically, it is not necessary to require that the circuit reach an exact equilibrium. Relatively small oscillations in the circuit are acceptable. We define a “relative steady state” where, within a time block of tblock = 200, the standard deviation of the time series of each node x(t) satisfies: std(x) < 1 × 10−4 and std(x)/mean(x) < 0.05. To further guarantee that the circuit is actually in the “relatively steady state,” two successive time blocks satisfying the standard deviation requirements are needed. The quantities O1 or O2 are then defined as the respective mean values of the output signal in that last time block tblock.

Relaxation time

An ideal gene circuit should be able to respond and adapt within a reasonable time scale. We set an upper bound of evolution time tmax = 4, 000. If the system cannot reach the “relative steady state” within this time, it is regarded as non-functional.

Details of parameter space sampling and response simulation

A three-node gene circuit subject to growth feedback has a large number of parameters. Let L be the number of links among the three nodes (excluding the input link). The total number of parameters is 2 · 3 + 2(L+ 1) = (2L+ 8). The values of these parameters determine the properties of the regulation links within the circuit and, as a result, the circuit dynamics. The circuit parameters are randomly generated by the Latin hypercube sampling method (Iman et al., 1980) using the function “lhsdesign” in Matlab. The parameters are sampled uniformly either on a logarithmic or a linear scale. The sampling ranges of the parameters are: νx ∈ [10−1, 101] (sampled in logarithmic scale), dx ∈ [10−2, 1] (sampled in logarithmic scale), nxy ∈ [1, 4] (sampled in linear scale), and Kxy ∈ [10−3, 1] (sampled in logarithmic scale).

The dynamical equations of the circuits are numerically integrated by the 4th order Runge–Kutta method with a time step Δt = 0.05. All the initial states of A, B, and C are taken to be 0.1. The input signal is initially I0 = 0.06 and then switched to I1 = 0.6.

We also observe that, when an isolated circuit fails, a certain amount of growth feedback can restore the circuit’s functions. This phenomenon was previously discovered experimentally (Tan et al., 2009). However, such cases are rare. We thus focus on circuits that are functional in isolation and examine how growth feedback affects their adaptation.

eLife Assessment

This valuable study focuses on the impact of growth feedback on the performance of artificial gene circuits capable of achieving adaptive responses, a significant problem in synthetic biology. Through solid computational analysis, the authors identify specific failure mechanisms, as well as core topologies associated with robust performance based on systematic analysis of over four hundred circuit topologies. The results will be of interest to those working on engineering gene circuits for diverse applications.

Joint Public Review:

Using computational modeling, this manuscript explores the effect of growth feedback on the performance of gene networks capable of adaptation. The authors selected 425 hypothetical synthetic circuits that were shown to achieve nearly perfect adaptation in two earlier computational studies (see Ma et al. 2009, and Shi et al. 2017). They examined the effects of cell growth feedback by introducing additional terms to the ordinary differential equation-based models, and performed numerical simulations to check the retainment and the loss of the adaptation responses of the circuits in the presence of growth feedback. The authors show that growth feedback can disrupt the gene network adaptation dynamics in different ways, and report some exceptional core motifs which allow for robust performance in the presence of growth feedback. They also used a metric to establish a scaling law between a circuit robustness measure and the strength of growth feedback. These results have important implications in the field of synthetic biology, where unforeseen interactions between designed gene circuits and the host often disrupt the desired behavior. The paper's conclusions are supported by their simulation results, although these are presented in their summary formats and it would be useful for the community if the detailed results for each topology were available as a supplementary file or through the authors' GitHub repository.

Strengths

- This work included a detailed investigation of the reasons for adaptation failure upon introducing cell growth to the systems. The comprehensiveness of the analysis makes the work stand out among studies of functional screening of network topologies of gene regulation.

- The authors' approaches for assessment of robustness, such as the survival ratio Q, can be useful for a wide range of topologies beyond adaptation. The scaling law obtained with those approaches is interesting.

Weaknesses

- The title suggests that the work investigates the 'effects of growth feedback on gene circuits'. However, the performance of 'nearly perfect adaptation' was chosen for the majority of the work, leaving the question of whether the authors' conclusion regarding the effects of growth feedback is applicable to other functional networks.

- This work relies extensively on an earlier study, evaluating only a selected set of 425 topologies that were shown to give adaptive responses (Shi et al., 2017). This limited selection has two potential issues. First, as the authors mentioned in the introduction, growth feedback can also induce emerging dynamics even without existing function-enabling gene circuits, as an example of the "effects of growth feedback on gene circuits". Limiting the investigation to only successful circuits for adaptation makes it unclear whether growth feedback can turn the circuits that failed to produce adaptation by themselves into adaptation-enabling circuits. Secondly, as the Shi et al. (2017) study also used numerical experiments to achieve their conclusions about successful topologies, it is unclear whether the numerical experiments in the present study are compatible with the earlier work regarding the choice of equation forms and ranges of parameter values. The authors also assumed that all readers have sufficient understanding of the 425 topologies and their derivation before reading this paper.

- The authors' model does not describe the impact of growth via a biological mechanism: they model growth as an additional dilution rate and calculate growth rate based on a phenomenological description with growth rate occurring at a maximum (k_g) scaled by the circuit 'burden' b(t). Therefore, the authors' model does not capture potential growth rate changes in parameter values (e.g., synthetic protein production falls with increasing growth rate; see Scott & Hwa, 2023).

- The authors made several claims about the bifurcations (infinite-period, saddle-node, etc) underlying the abrupt changes leading to failures of adaptations. There is a lack of evidence supporting these claims. Both local and global bifurcations can be demonstrated with semi-analytic approaches such as numerical continuation along with investigations of eigenvalues of the Jacobian matrix. The claims based on ODE solutions alone are not sound.

- The impact of biochemical noise is not evaluated in this work; the author's analysis is only carried out in a deterministic regime.

Author response:

The following is the authors’ response to the original reviews.

Point-by-point response to the public review:

General Comment: “Using computational modeling, this manuscript explores the effect of growth feedback on the performance of gene networks capable of adaptation. The authors selected 425 hypothetical synthetic circuits that were shown to achieve nearly perfect adaptation in two earlier computational studies (see Ma et al. 2009, and Shi et al. 2017). They examined the effects of cell growth feedback by introducing additional terms to the ordinary differential equation-based models, and performed numerical simulations to check the retainment and the loss of the adaptation responses of the circuits in the presence of growth feedback. The authors show that growth feedback can disrupt the gene network adaptation dynamics in different ways, and report some exceptional core motifs which allow for robust performance in the presence of growth feedback. They also used a metric to establish a scaling law between a circuit robustness measure and the strength of growth feedback. These results have important implications in the field of synthetic biology, where unforeseen interactions between designed gene circuits and the host often disrupt the desired behavior. The paper’s conclusions are supported by their simulation results, although these are presented in their summary formats and it would be useful for the community if the detailed results for each topology were available as a supplementary file or through the authors’ GitHub repository.”

We are grateful for the referee’s positive evaluation of our work. We have updated our GitHub and OSF repositories with detailed results for each topology. Additionally, we have included other simulation codes, result data, and detailed explanations in these two repositories that may be of interest to our readers.

Strength 1: “This work included a detailed investigation of the reasons for adaptation failure upon introducing cell growth to the systems. The comprehensiveness of the analysis makes the work stand out among studies of functional screening of network topologies of gene regulation.”

We are grateful for the referee’s positive assessment of our work, notably the recognition of the ‘detailed investigation’ we conducted, and the ‘comprehensiveness of the analysis’ we provided.

Strength 2: “The authors’ approaches for assessment of robustness, such as the survival ratio Q, can be useful for a wide range of topologies beyond adaptation. The scaling law obtained with those approaches is interesting.”

We are grateful for the referee’s positive evaluation of our defined factors for assessing circuit robustness. We also appreciate the acknowledgment of the “interesting” nature of the scaling law we discovered using the assessment factor R.

Weaknesses 1: “The title suggests that the work investigates the ’effects of growth feedback on gene circuits’. However, the performance of ’nearly perfect adaptation’ was chosen for the majority of the work, leaving the question of whether the authors’ conclusion regarding the effects of growth feedback is applicable to other functional networks.”

We agree that our present title can be too broad, and we have changed it from “Effects of growth feedback on gene circuits: A dynamical understanding” to “Effects of growth feedback on adaptive gene circuits: A dynamical understanding”. Although we have some brief results and discussions on the gene circuits with bistability, we admit that most of our results and discussions are focused on circuits that have adaptation.

The new title is more specific and should be a more appropriate summary of the paper.

Weaknesses 2: “This work relies extensively on an earlier study, evaluating only a selected set of 425 topologies that were shown to give adaptive responses (Shi et al., 2017). This limited selection has two potential issues. First, as the authors mentioned in the introduction, growth feedback can also induce emerging dynamics even without existing function-enabling gene circuits, as an example of the ”effects of growth feedback on gene circuits”. Limiting the investigation to only successful circuits for adaptation makes it unclear whether growth feedback can turn the circuits that failed to produce adaptation by themselves into adaptation-enabling circuits. Secondly, as the Shi et al. (2017) study also used numerical experiments to achieve their conclusions about successful topologies, it is unclear whether the numerical experiments in the present study are compatible with the earlier work regarding the choice of equation forms and ranges of parameter values. The authors also assumed that all readers have sufficient understanding of the 425 topologies and their derivation before reading this paper.”

We agree with the reviewer that several issues need to be clarified in our new manuscript. We have added new discussions for all of them.

We agree with the reviewer that growth feedback could turn the non-adaptive circuits into adaptationenabling circuits, and this indeed presents a compelling topic for future research. We have added the following discussions to our paper, talking about a relevant matter. We find that in our simulated dataset, there are cases where a higher degree of growth feedback can restore the adaptation that has been lost in a circuit. However, as we discussed in this new paragraph, a comprehensive study in the direction of turning non-adaptive circuits into adaptation-enabling circuits will “require entirely different approaches for sampling circuit parameters and selecting candidate network topologies, demanding significantly high computational costs.” Given that this topic extends beyond the scope of the current paper, we leave this matter to future research.

“Although the primary focus of this paper is on how growth feedback can undermine an originally adaptive circuit and how to design circuits that are robust against such feedback, our simulated dataset reveals instances where growth feedback can benefit the circuit within certain ranges. Specifically, we identified 2,092 circuits across 306 different topologies where adaption, lost at an intermediate level of growth feedback, is restored at higher levels. This is 1.4% of all circuits tested. We anticipate that additional circuits exhibiting this loss-and-recovery behavior exist, as our sampling of six discrete levels of kg (0,0.2,0.4,0.6,0.8,1.0) might have overlooked numerous cases. This result again suggests the possible advantages of growth feedback in gene circuits (Tan et al., 2009; Nevozhay et al., 2012; Deris et al., 2013; Feng et al., 2014; Melendez-Alvarez and Tian, 2022). A comprehensive study into how growth feedback can endow or enhance adaption in circuits would require entirely different approaches for sampling circuit parameters and selecting candidate network topologies, demanding significantly high computational costs. Given that this topic extends beyond the scope of the current paper, we leave this matter to future research.”

We have added the following discussions about the reasoning behind using the 425 network topologies selected from the study Shi et al. (2017).

“We use these 425 network topologies from the study (Shi et al., 2017), avoiding redundancy with established results. Due to the unique focus of our research on the effects of growth feedback and the need to evaluate quantitative ratios of robust circuits among all functional ones, we have chosen to use a 20-fold increase in the number of random parameter sets for each network topology compared to the simulations in (Shi et al., 2017). This approach makes it computationally prohibitive to scan all possible 16,038 three-node circuits. We carefully follow the settings in (Shi et al., 2017), which also analyzed TRNs with the AND logic as in this paper. Detailed descriptions of our simulation experiments are provided in the Methods section. To make our results more convincing, we have adopted a set of adaptation criteria that are stricter than those used in (Shi et al., 2017). Consequently, the ratio of adaptive circuits is somewhat lower in our study, with 4 out of the 425 network topologies not demonstrating adaptation.”

Other than the more strict adaptation criteria and much larger sampling sizes, as we mentioned in this paragraph, we have carefully followed the simulation details of the study Shi et al. (2017). This includes but is not limited to: the dynamical equations (when kg = 0), the input signals, the scales and ranges of the circuit parameters to be randomly sampled, and the sampling method (Latin hypercube sampling). One of the authors of the current paper was also the first author of the study Shi et al. (2017), who helped us verify the details of simulations (among many other contributions). These identical settings justify our usage of the established results with the 425 network topologies.

To provide more information about these 425 network topologies, We have added the following introduction. It introduces the structural features of the networks, especially the shared core motifs for adaptation. In our GitHub and OSF repositories, we have also provided relevant data about the 425 topologies, including the topology structures and the parameter sets we scanned.

“These topologies can be classified into two families based on the core topology: networks with a negative feedback loop (NFBL) and networks with an incoherent feed-forward loop (IFFL) (Shi et al., 2017). More specifically, there are 206 network topologies in the NFBL family. All of these NFBL topologies have a negative feedback loop for node B. This negative feedback loop can be formed by the loop from node B to A and back to B (such as the circuit shown in Fig. 1 (a)), by node B to C and back to B, or by a longer route, from node B to A and then to C and back to B. There is always a self-activation link from B to B in all these 206 NFBL networks. There are 219 network topologies in the IFFL family. All of them have two feed-forward pathways from the input node A to the output node C. One pathway goes from node A to C directly, while the other involves node B in the middle. One of the pathways is activating while the other one is inhibitory.”

Weaknesses 3: “The authors’ model does not describe the impact of growth via a biological mechanism: they model growth as an additional dilution rate and calculate growth rate based on a phenomenological description with growth rate occurring at a maximum (kg) scaled by the circuit ’burden’ b(t). Therefore, the authors’ model does not capture potential growth rate changes in parameter values (e.g., synthetic protein production falls with increasing growth rate; see Scott & Hwa, 2023).”

In our paper, we consider dilution due to cell growth as the dominant factor of growth feedback. Here we compared the adaptive circuits under no-growth conditions and their ability to maintain their adaptive behaviors after dilution into a fresh medium, which mediated a significant dilution to the circuits. This is based on our previous work, Zhang, et al. Nature chemical biology 16.6 (2020): 695-701. We agree that an increased growth rate can change synthetic protein production. However, the dynamic roles of the dilution and growthaffected production rate should be analogous, given that they both act as inhibitory factors arising from cell growth as mentioned by the reviewer. Still, we agree that taking the growth effect on the production rate into account would provide a more comprehensive study, but it is beyond the scope of the present work. We have added the following paragraph in the Discussion section of our paper.

“In our paper, we consider dilution due to cell growth as the dominant factor of growth feedback. Here we compared the adaptive circuits under no-growth conditions and their ability to maintain their adaptive behaviors after dilution into a fresh medium, which mediated a significant dilution to the circuits. This is based on our previous work (Zhang et al. (2020)). However, growth feedback is inherently complex (Klumpp et al. (2009)). For instance, an increased growth rate can change protein synthesis rate (Hintsche and Klumpp (2013); Scott and Hwa (2023)), and cell growth rates can affect the distribution of protein expression in cell populations (Gouda et al. (2019)). In our paper, we concentrate on a simplified model with dilution, which we consider to have captured the dominant factor. The dynamic roles of the dilution and growth-affected production rate should be analogous, given that they both act as inhibitory factors arising from cell growth. Incorporating the impact of growth rate on protein synthesis into our model would offer a more comprehensive analysis, a task beyond the scope of this paper but presenting an intriguing opportunity for future research to address the complexities of growth feedback.”

Weaknesses 4: “The authors made several claims about the bifurcations (infinite-period, saddle-node, etc) underlying the abrupt changes leading to failures of adaptations. There is a lack of evidence supporting these claims. Both local and global bifurcations can be demonstrated with semi-analytic approaches such as numerical continuation along with investigations of eigenvalues of the Jacobian matrix. The claims based on ODE solutions alone are not sound.”

After our further simulations and verification, we found that most of the bifurcation-induced failures we mentioned in type-V and type-VI failures should be categorized as bistability or multistability-induced failures. They are still abrupt switching between adaptive and non-adaptive states, as we described in the previous version of the manuscript. However, they are actually still far away from the bifurcation points at the critical kg. We have corrected all relevant descriptions and figures, including panel Fig. 4 (c) and its captions. We have added the following paragraph in the paper to explain this issue.

“One might expect bifurcations to play an important role in many type-V and type-VI failures. However, in our simulations, failures precisely at the bifurcation point are not observed. This is because the bifurcation points under consideration, such as fold bifurcations, are where one of the attraction basins diminishes to zero. For a failure to occur exactly at the bifurcation point, the initial condition would need to coincide precisely with the infinitesimally small basin just before it vanishes. More realistically, failures almost always largely precede the exact bifurcation point. They happen while the basin is still contracting and the basin boundary crosses the initial condition or O1. An example is shown in Fig. 4(b), where bistability persists, yet the lighter orange basin with a larger O1(C) cannot be reached as the boundary shifts away from the initial condition A0 and B0. As another example, in Fig. 4 (c) from a different circuit, the higher O2(C) state disappears at kg ≈ 0.012 and switches to a lower O2(C), but this point is not a bifurcation.

It is the point where the stable O1 continuously crosses the basin boundary of O2.”

Our further simulations have verified the existence of the oscillation-related bifurcations. We have added a new appendix discussing the phenomena associated with them in more detail.

Weaknesses 5: “The impact of biochemical noise is not evaluated in this work; the author’s analysis is only carried out in a deterministic regime.”

In this paper, we have not taken into account biochemical noise as we focus solely on scenarios where all protein concentrations are high. In these circumstances, the influence of noise is relatively minor. Incorporating biochemical noise, which originates from various sources and possesses diverse characteristics, would significantly complicate the analysis beyond the scope of our current work. However, exploring this aspect could be an intriguing avenue for future research. We have included the following discussions in our paper.

“Our study focuses on scenarios where random noises are ignored. Realistically, gene circuits are subjected to diverse types of noise, which can complicate their predictable behavior and design. These noises can originate externally from a noisy input signal I, or intrinsically, directly affecting the circuit components. Further, these noises can be classified based on various mechanisms that cause them (Colin et al. (2017); Sartori and Tu (2011)) . And with different mechanisms, each type of noise can be characterized by different attributes such as frequency, amplitude, and noise color. These variances can lead to different impacts on the circuits, potentially necessitating unique mechanisms or designs for the attenuation of each category (Sartori and Tu (2011); Qiao et al. (2019) ). Given the extensive complexity and the need for thorough investigation, these noise-related challenges are beyond the scope of this paper and require a series of future studies.”

Point-by-point response to the recommendations for the authors:

Comment 1: - The authors’ github repository, detailed in their code availability statement, is currently unavailable and likely contains some of the answers to the queries here.

We have updated our GitHub and OSF repositories with simulation codes, result data, and detailed explanations. The link to our GitHub repository in the previous version of the manuscript contained a format error, making it inaccessible to the referees. We apologize for this mistake and have corrected it.

Comment 2: - At present, it is not clear how the 425 topologies are created from the system of equations (Eq. 6-8) or from the circuit diagram in Fig 1a. This could do with being explicitly stated for the reader.

We have added the following paragraph to discuss how the 425 topologies are selected and what the common motifs and connections they share.

“Previous research identified 425 different three-node TRN network topologies that can achieve adaptation in the absence of growth feedback (Shi et al., 2017), providing the base of our computational study. These topologies can be classified into two families based on the core topology: networks with a negative feedback loop (NFBL) and networks with an incoherent feed-forward loop (IFFL) (Shi et al., 2017). More specifically, there are 206 network topologies in the NFBL family. All of these NFBL topologies have a negative feedback loop for node B. This negative feedback loop can be formed by the loop from node B to A and back to B (such as the circuit shown in Fig. 1 (a)), by node B to C and back to B, or by a longer route, from node B to A and then to C and back to B. There is always a self-activation link from B to B in all these 206 NFBL networks. There are 219 network topologies in the IFFL family. All of them have two feed-forward pathways from the input node A to the output node C. One pathway goes from node A to C directly, while the other involves node B in the middle. One of the pathways is activating while the other one is inhibitory. We use these 425 network topologies from the study (Shi et al., 2017), avoiding redundancy with established results. Due to the unique focus of our research on the effects of growth feedback and the need to evaluate quantitative ratios of robust circuits among all functional ones, we have chosen to use a 20-fold increase in the number of random parameter sets for each network topology compared to the simulations in (Shi et al., 2017). This approach makes it computationally prohibitive to scan all possible 16,038 three-node circuits. We carefully follow the settings in (Shi et al., 2017), which also analyzed TRNs with the AND logic as in this paper. Detailed descriptions of our simulation experiments are provided in the Methods section. To make our results more convincing, we have adopted a set of adaptation criteria that are stricter than those used in (Shi et al., 2017). Consequently, the ratio of adaptive circuits is somewhat lower in our study, with 4 out of the 425 network topologies not demonstrating adaptation.”

Comment 3: - In the main text, the authors mentioned that they chose 425 network topologies for this study, whereas the number is 435 in the abstract. Please correct the error.

The number 435 in our previous abstract referred to the 10 four-node circuits that we studied in the appendix, in addition to the 425 three-node network topologies. To avoid confusion and potential misunderstandings among readers, we have revised this expression of “435 distinct topological structures” to “more than four hundred topological structures”.

Comment 4: - Please can the authors include the topologies they have studied in an appendix or as supplementary material. The impact of this work would increase significantly if for each topology the authors could include a pie chart similar to the one shown in Fig 2 so that others can use these results.

We fully acknowledge the potential benefits of providing simulation results for each topology. However, including over four hundred more figures in this paper is not feasible. Moreover, we expect that many readers may also be interested in results not only for individual topologies but also for subsets sharing specific motifs or regulatory connections. Therefore, we have provided all the necessary data and codes in our GitHub repository to make these pie charts. We have included a detailed guide on how to generate these pie charts in the GitHub Readme file. These allow readers to plot the pie chart and extract distributions for any individual topology or use conditions to filter any subset of topologies as required. We believe this approach offers greater flexibility for our readers. We have also added the following explanation in the Methods section.

“The codes implementing these criteria are available in our GitHub repository, with the link provided in the ”Code Availability” section. The failure type results for all circuits tested are available in our OSF repository, with the link provided in the ”Data Availability” section. An additional note is provided in the README file of our GitHub repository for further guidance on generating pie charts similar to Fig. 2 for any network topology or subset of topologies.”

Comment 5: - At present, the authors have not given sufficient detail for their numerical methods (e.g. to identify bistability or oscillations) to enable the work to be repeated. I would appreciate it if the authors could expand their Methods section or provide a description of their method as an appendix. Additionally, the authors must clarify how many parameter sets per topology showed successful adaptation.

In response to this comment, we have reorganized and expanded our Methods section, especially the new “Numerical simulations of circuit dynamics” and “Numerical criteria for functional adaptation and failure types” subsections. We added details on how we define and evaluate a “relatively steady state”, how to determine if there is an oscillation, how to determine the critical kg value, and how to determine if a failure is continuous or abrupt. Readers can also find the corresponding codes in our GitHub repository, where we provide a README file to help the readers locate the script file they need.

The number of parameter sets per topology showed successful adaptation is precisely our definition of the Q-value. Q-values of most of the circuits we tested are shown in multiple figures in the paper. A complete table of Q-values with different topologies and different kgrowth values can be found in our OSF repository.

Comment 6: - Looking at the Model Description, there seem to be multiple issues, as follows. The model should be rewritten and all simulations redone with the model corrected as described below:

(a) The ”strength of growth feedback” is modeled by the maximal growth parameter kg in Equation (12). However, this rate does not represent growth feedback. In fact, this parameter must be present also for the system without growth feedback, Equations (6 - 8), because those cells grow as well! So Equation (12) with b(t)=0 should also be added to Equations (6 - 8), in addition to the dilution terms in each equation.

(b) The dilution due to growth (dN/dt)*(B/N) is only added to Equations (9 - 11). This is wrong - growthaffects (dilutes) all protein concentrations, even without growth feedback, so similar terms must be added even to equations without growth feedback, i.e., to Equations (6 - 8).

(c) The term representing growth feedback is actually the fraction 1/(1+b(t)). To adjust the strength ofgrowth feedback, some parameters should be introduced into this term. Specifically, the term currently has a Hill form with Hill coefficient = 1 and sensitivity = 1. The term should be converted into a general Hill function, and the parameters of that function should be altered to represent growth feedback. This Hill function is called a cellular (phenotypic) fitness landscape, see Nevozhay et al., 2012.

Equations (6-8) only describe one part of the entire model we are studying. We are having these equations presented solely for the purpose of not overwhelming readers with a large number of parameters that are defined for the first time. They are not actually used in our simulations, but were only for explanations of the meaning of parameters. In our simulations throughout the paper, we only used Eqs. (9-13) (with various topologies). We have revised the texts to make this point clear. We have added the following descriptions in the section Model Description:

“In order not to overwhelm readers with too many terms and parameters, we first describe a partial model (an isolated circuit without growth feedback) before introducing the complete model that we study in this work.”

“Equations. (9) to (13) are the dynamical equations we actually use for simulating the circuit dynamics.”

Additionaly, in the newly added subsection “Numerical simulations of circuit dynamics683” in the Methods, we explicitly mention that:

“The dynamical equations we use are similar to Eqs. (9-13) but with different topologies.”

We consider dilution due to cell growth as the dominant factor of growth feedback. In fact, we study the adaptive circuits without growth and their ability to maintain their adaptive behaviors after dilution into a fresh medium, based on a recent work [Zhang, et al., Nature Chemical Biology 16.6 (2020): 695-701]. The dynamic roles of the dilution and growth-affected production rate should be analogous, given that they both act as inhibitory factors arising from cell growth. The term mentioned in the comment is about how the burden of the circuit affects cell growth. We agree that it can be interesting to have a more comprehensive study on how different degrees of nonlinearity of this term can have different effects on the overall robustness towards the growth feedback problem, but this is not part of our primary focus and is beyond the scope of this paper. In this paper, we are mostly concerned with the variability of the strength of the growth feedback/dilution, controlled by the parameter kg, instead of the different types of nonlinearity.

Comment 7: - On the right side of Equation (7), the first term should be inhibitory, right?

This is indeed an error. We accidentally reversed the regulation from A to B and B to A when inputting the formula. We have corrected both terms.

Comment 8: - It seems to me that a better transition from Figs 6 and 7 to Fig 8 can be made. Did the authors choose the three circuits in Fig 8 based on the three distinct groups shown in Fig 6 and 7? The rationale for choosing the three topologies given the clusters identified earlier can be explained more clearly.

We agree more explanation can be provided here. We have added the following descriptions, in the caption of Fig.8:

“The other three curves represent circuits with different robustness levels: high (Circuit No. 98), moderate (Circuit No. 3), and low (Circuit No. 28) values of R, to demonstrate that this scaling behavior is generic. Each of these three circuit topologies is selected from one of the three groups illustrated in Fig. 6 and Fig. 7, and they have the highest Q(kg = 0) value within their respective groups.”

and in the main text:

“The three other curves represent circuit topologies that have a relatively high, moderate, and low value R among the 425 topologies tested, to demonstrate that this scaling behavior is generic. (These three topologies are the highest Q(kg = 0) topology in each of the three groups shown in Fig. 6 and Fig. 7.”

Comment 9: - The insights from the neural network model seem to be very limited. It would be interesting to see if the model can predict the performance of network topologies that have not been exposed to the model during training.

Machine learning is not a focus of this paper. For the section the comment was referring to, the main research question is on the relationship between circuit robustness and topology, and the point we are trying to make is that the robustness dependency varies across different connections — some connections are critical, while others are less impactful. The neural-network-based analysis was only used to provide further support to this point by demonstrating that through optimization, neural networks automatically assign different levels of weights to different connections in the circuits.

We agree that it can be an interesting topic to study how machine learning can be used to help us design functional and robust circuits, as discussed in the final paragraph of the Discussion section. However, such an investigation would require a series of more comprehensive and carefully designed simulation experiments to validate if “neural networks can predict the performance of network topologies that have not been exposed to the model during training”. One point one should take extra care of is that many network topologies we study are very similar to many others, with shared motifs and links. These considerations extend beyond the scope of this paper.

Other potential improvements or future work

Comment 10: - The growth feedback examined in this paper comes from the effect of protein levels on the cell division rate (growth rate). However, the opposite effect can also occur; cell growth rates can affect the distribution of protein expression in cell populations. A good reference is Kheir Gouda et al., which is already on the list of references. These opposite effects should be described and discussed.

We agree that growth feedback is inherently complex and has many biological effects, and in our paper, we are using a simplified model to study the dominant factor of growth feedback. We have added the following paragraph in the Discussion section, which involves the opposite effect mentioned in the comment.

“In our paper, we consider dilution due to cell growth as the dominant factor of growth feedback. Here we compared the adaptive circuits under no-growth conditions and their ability to maintain their adaptive behaviors after dilution into a fresh medium, which mediated a significant dilution to the circuits. This is based on our previous work (Zhang et al. (2020)). However, growth feedback is inherently complex (Klumpp et al. (2009)). For instance, an increased growth rate can change protein synthesis rate (Hintsche and Klumpp (2013); Scott and Hwa (2023)), and cell growth rates can affect the distribution of protein expression in cell populations (Gouda et al. (2019)). In our paper, we concentrate on a simplified model with dilution, which we consider to have captured the dominant factor. The dynamic roles of the dilution and growth-affected production rate should be analogous, given that they both act as inhibitory factors arising from cell growth. Incorporating the impact of growth rate on protein synthesis into our model would offer a more comprehensive analysis, a task beyond the scope of this paper but presenting an intriguing opportunity for future research to address the complexities of growth feedback.”

Comment11: - It may be worth mentioning that growth feedback can lead to persistence, see PMID:27010473.

We have included this research as a citation.

Comment 12: - While some other networks (two-node) are discussed, it would be worth doing this analysis for all one- and two-node networks, perhaps controlled by small molecules added externally. If not here, then as a future plan.

We agree that this is an interesting idea for future studies.

Comment 13: - The manuscript analyzes the deterministic dynamics of a set of gene networks. However, gene expression is always stochastic, and gene circuits have been designed to control stochastic gene expression. For example, gene expression distributions can be reshaped, or even new peaks can appear, which would be worth mentioning, PMID: 30341217. The effect of growth feedback on stochastic gene expression and future perspectives of systematically studying this should be discussed.

We have added the following paragraph in the Discussion section to discuss the effects of noises and stochasticity. The research mentioned in the comment is also included.

“Our study focuses on scenarios where random noises are ignored. Realistically, gene circuits are subjected to diverse types of noise, which can complicate their predictable behavior and design. These noises can originate externally from a noisy input signal I, or intrinsically, directly affecting the circuit components. Further, these noises can be classified based on various mechanisms that cause them (Colin et al. (2017); Sartori and Tu (2011)). And with different mechanisms, each type of noise can be characterized by different attributes such as frequency, amplitude, and noise color. These variances can lead to different impacts on the circuits, potentially necessitating unique mechanisms or designs for the attenuation of each category (Sartori and Tu (2011); Qiao et al. (2019)). Given the extensive complexity and the need for thorough investigation, these noise-related challenges are beyond the scope of this paper and require a series of future studies.”

Version 2

Abstract

Abstract

The successful integration of engineered gene circuits into host cells remains a significant challenge in synthetic biology due to circuit-host interactions, such as growth feedback, where the circuit influences cell growth and vice versa. Understanding the dynamics of circuit failures and identifying topologies resilient to growth feedback are crucial for both fundamental and applied research. Utilizing transcriptional regulation circuits with adaptation as a paradigm, we systematically study more than four hundred topological structures and uncover various categories of failures. Three dynamical mechanisms of circuit failures are identified: continuous deformation of the response curve, strengthened or induced oscillations, and sudden switching to coexisting attractors. Our extensive computations also uncover a scaling law between a circuit robustness measure and the strength of growth feedback. Despite the negative effects of growth feedback on the majority of circuit topologies, we identify several circuits that maintain optimal performance as designed, a feature important for applications.

Introduction

In biomedical science and engineering, artificially designed gene circuits are anticipated to play an ever-increasing role in disease diagnosis and therapy (Riglar and Silver, 2018; Sedighi et al., 2019; Xia et al., 2019). Gene circuits also show great potential in various applications such as microbiome modulation (Foo et al., 2017; Lee et al., 2018) and biological containment (Gomaa et al., 2014; Caliando and Voigt, 2015). While most gene circuits are designed to function after they are inserted or embedded into host cells, the interactions between the circuit and the host environment are generally extremely complex and can lead to undesired effects that were not present in the original, isolated circuit (Tan et al., 2009; Ceroni et al., 2015; Borkowski et al., 2016; Ceroni et al., 2018; Darlington et al., 2018a,b; Gouda et al., 2019; Zhang et al., 2021, 2020; Melendez-Alvarez et al., 2021). Understanding the interactions and identifying the circuit topological structures that can withstand the interactions and thrive in the host are thus of fundamental importance, requiring interdisciplinary efforts among systems and synthetic biology, metabolic engineering, nonlinear dynamics and complex systems.

Typical circuit-host interactions include metabolic burden, cell growth, and resource relocation or competition, among which growth feedback is the most common type of circuit-host interaction between the circuit gene expressions and cell growth. More specifically, a synthetic gene circuit embedded in a host cell possesses an intrinsic coupling mechanism: the circuit affects cell growth and the growth in turn modifies the gene expressions in the circuit (Klumpp et al., 2009; Klumpp and Hwa, 2014; Boo et al., 2019; Scott et al., 2010; Ray et al., 2016) - the so-called growth feedback. Studies have shown that the growth-mediated feedback can endow a synthetic gene circuit with various emergent properties, such as innate growth bistability (Deris et al., 2013). For example, a non-cooperative positive autoregulation system, when coupled with growth feedback, gains increased effective cooperativity, thereby resulting in bistability (Tan et al., 2009; Nevozhay et al., 2012). In another example, toxin cooperativity can be induced in multiple toxin-antitoxin systems by growth-mediated feedback (Feng et al., 2014). The number of steady states in one gene circuit also depends on growth feedback and resource availability (McBride and Del Vecchio, 2020; Melendez-Alvarez and Tian, 2022). In general, growth feedback acts to hamper the forward engineering of the circuit functions by introducing modes of nonmodularity and reducing the predictability of the circuit components in an in-vivo context. While various phenomena caused by growth feedback were studied with desirable or undesirable effects on the functions of the gene circuits, a systemic picture is lacking on what effects growth feedback can have on the gene circuits, including failures.

A recent study has revealed that growth feedback can have drastically different effects on congruent circuits with distinct topologies (Zhang et al., 2020). In particular, the dynamical behaviors of two bistable synthetic memory circuits were studied: a self-activation switch incorporating positive autoregulation and a toggle switch incorporating double-negative regulatory motifs. It was found that growth feedback impacts both circuits but with quite different manifestations. For the toggle switch, memory can be retained and the circuit tends to be refractory towards growth feedback. However, for the self-activation switch, growth feedback leads to memory loss. While these results indicate that the circuit topology can play a significant role in the circuit functions when growth feedback is present, they were obtained through two specific circuit topologies. Since a particular function of the gene circuit can often be achieved by a finite set of core topologies, it is of fundamental interest to identify the most robust topologies in response to growth feedback. The so-identified optimal topologies can then be used to construct synthetic gene circuits capable of maintaining the essential functions to meet the design goals under the fluctuating growth conditions of the host cell. A systematic study of the interplay between the gene circuit topology and growth feedback is needed.

Adaptation is an important and widely studied functionality of gene circuits, which is defined as the ability of the system to respond to environmental changes and to return to the basal or nearbasal state after some time (Knox et al., 1986; Tyson et al., 2003; Friedlander and Brenner, 2009; Ferrell Jr, 2016). Previously, it was found that certain circuits possess biochemical adaptation (Ma et al., 2009) if they contain at least one of the two architectural classes: an incoherent feed-forward loop with a proportion node and a negative feedback loop with a buffering node. A number of synthetic gene circuits were proposed or constructed to achieve adaptation (Kim et al., 2014; Briat et al., 2016; Aoki et al., 2019). Quite recently, a design principle for circuits with four genes was uncovered for simultaneously achieving noise attenuation and adaptation: the circuit must have a sequential assembly structure (Qiao et al., 2019). However, these existing adaptation studies did not include any growth feedback mechanism.

In this paper, we conduct a comprehensive computational study to uncover and understand the effects of growth feedback on the gene circuits. Specifically, we focus on a type of transcriptional regulation circuit designed for adaptation. There are 425 possible circuit structures (identified by previous research (Shi et al., 2017)), and we study all of them to simulate and test their response under different levels of growth feedback. Altogether, 2 × 105 sets of circuit parameters are randomly sampled for each structure. Our results reveal a vast number of cases where growth feedback has a detrimental effect on circuit function (1.3 × 105 cases in total) with varying response curves and dynamical behaviors. To gain a more intuitive overall picture, we classify these cases into several distinct categories based on the circuits’ dynamic behavior. We then systemically summarize the dynamical mechanism behind these growth-induced circuit malfunctions. To quantify circuit adaptation in the presence of growth feedback, we propose a robustness measure that enables us to identify an optimal group of circuits that exhibit a high level of robustness against growth feedback, making them particularly promising for real-world implementation. The motifs associated with this optimal group are found through machine learning. We also obtain a scaling law governing the dependence of this measure on the level of growth feedback and provide a mathematical analysis to gain insights into the underpinnings of the scaling law. The take-home message is that, in spite of the negative effects of growth feedback in the majority of the circuits, there exists a small set of circuits that are still able to deliver optimal performance as designed, which is promising for real-world implementation.

Results

A systemic search of functional failures due to growth feedback

Adaptation is referred to as the ability of a gene circuit to respond to changes in input and then to return to the pre-stimulus output level, even when the input change persists (Ma et al., 2009). More precisely, with an input signal switched from a lower value I1 to a higher value I2, as demonstrated in Fig. 1(b), a circuit with functional adaptation should have the following response-curve criteria: (i) precision - the final state O2 should be close to the initial state O1, (ii) sensitivity - there should be a relatively high |Opeak| in response to the change in the input, and (iii) the system should reach equilibrium within a reasonable relaxation time. A threenode gene circuit can achieve adaptation (Ma et al., 2009), with one node receiving the input (node A), another node realizing various regulatory roles (node B), and a third node outputting the response (node C). A representative circuit topology is shown inside the red dashed box in Fig. 1(a). We restrict our study of the class of transcriptional regulatory networks (TRNs) with the AND logic. We fix node A as the input node, and node C as the output node.

Schematic illustration of a synthetic gene circuit embedded in a host cell.

(a) A representative three-gene circuit (inside the dashed red box) and its dynamical interplay with host-cell growth. Arrows with triangular ends and round ends denote activating and inhibiting regulations, respectively. Altogether, there are 16,038 possible three-node topologies, with 425 topologies capable of adaptation. (b) An example of the circuit input and output signals. The input is an idealized step function of currents I1 and I2 before and after the jump, respectively. The output signal is a response of the circuit to the step function. The features of the output signal, as characterized by three key quantities characterizing the signal: O1, O2, and Opeak, can be used to determine if the circuit has succeeded or failed in its intended function.

Previous research identified 425 different three-node TRN network topologies that can achieve adaptation in the absence of growth feedback (Shi et al., 2017), providing the base of our computational study. These topologies can be classified into two families based on the core topology: networks with a negative feedback loop (NFBL) and networks with an incoherent feed-forward loop (IFFL) (Shi et al., 2017). More specifically, there are 206 network topologies in the NFBL family. All of these NFBL topologies have a negative feedback loop for node B.

This negative feedback loop can be formed by the loop from node B to A and back to B (such as the circuit shown in Fig. 1 (a)), by node B to C and back to B, or by a longer route, from node B to A and then to C and back to B. There is always a self-activation link from B to B in all these 206 NFBL networks. There are 219 network topologies in the IFFL family. All of them have two feedforward pathways from the input node A to the output node C. One pathway goes from node A to C directly, while the other involves node B in the middle. One of the pathways is activating while the other one is inhibitory. We use these 425 network topologies from the study Shi et al. (2017), avoiding redundancy with established results. Due to the unique focus of our research on the effects of growth feedback and the need to evaluate quantitative ratios of robust circuits among all functional ones, we have chosen to use a 20-fold increase in the number of random parameter sets for each network topology compared to the simulations in Shi et al. (2017). This approach makes it computationally prohibitive to scan all possible 16,038 three-node circuits. We carefully follow the settings in Shi et al. (2017), which also analyzed TRNs with the AND logic as in this paper. Detailed descriptions of our simulation experiments are provided in the Methods section. To make our results more convincing, we have adopted a set of adaptation criteria that are stricter than those used in Shi et al. (2017). Consequently, the ratio of adaptive circuits is somewhat lower in our study, with 4 out of the 425 network topologies not demonstrating adaptation. The specific structures of these 425 network topologies can be found in our GitHub repository (link provided in Data availability).

In our work, we use a parameter kg to control the strength of growth feedback, which is a parameter determining the maximal growth rate of the host cells, as mathematically explained in Model description. With all the other parameters fixed, a larger kg implies a faster cell growth rate and a stronger impact of growth feedback. To investigate the effect of growth feedback on these circuits, we systematically simulate the response of the 425 network topologies under a switch in the input signal. A three-node gene circuit subject to growth feedback has a large number of parameters, which determine the properties of the regulation links within the circuit and the circuit dynamics. For each topology, we randomly sample 2 × 105 trials of circuit parameters. Altogether, our study involves analyzing approximately 8.5 × 107 different circuits. We find that among these trials, only about 1.5 × 105 meet the adaptation criterion in the absence of growth feedback. For these functional trials, we vary the growth feedback parameter kg with a series of values, and find that the majority of trials (1.3×105 trials, about 87%) lose their adaptation in the interval of kg ∈ (0, 1), while only 13% of trials remain functional at kg = 1.0.

A systemic classification of functional failures due to growth feedback

An essential step towards understanding the detrimental or even destructive effects of growth feedback on circuit functioning is to identify the distinct failure scenarios. Our extensive simulations have yielded a comprehensive picture of these scenarios, as shown in Fig. 2. Overall, we have identified six failure scenarios that encompass more than 99.6% of the 1.3 × 105 failing cases we collect. The first level of classification distinguishes between failures that occur continuously or abruptly as the growth-feedback strength kg increases. In a continuous failure, the response curve deforms continuously as kg increases, as exemplified in Figs. 2(a-c). In an abrupt failure, the response curve exhibits a sudden change as kg increases through a critical value, as illustrated in Figs. 2(d-f). At the next classification level, we further divide the failures into three types of continuous failures and three types of abrupt failures.

Systemic classification of circuit failure scenarios due to growth feedback.

This study identifies six computationally detectable categories of failures based on the criterion of functional adaptation that the circuit violates as the effect of growth feedback becomes stronger. (a) Type-I and (b) type-II failures correspond to the cases where the precision criterion or sensitivity criterion is violated in a continuous fashion as the growth-feedback strength kg increases, respectively. (c) Type-III and (d) type-IV failures occur when the circuits lose adaptation due to growth-feedback-induced oscillation, either continuously or abruptly, as kg increases, respectively. The abrupt changes in type-IV are caused by bifurcations, mostly a saddle-node bifurcation of cycles or an infinite-period bifurcation. For instance, the case shown in (d) undergoes an infinite-period bifurcation. (e) Type-V and (f) type-VI failures are when the circuits lose adaptation due to an abrupt change in O1 or O2 as kg increases, respectively, which are caused by bistability or multistability in the systems. Trials that are not categorized under these six classifications or fall into multiple categories constitute less than 0.4% of all cases (see text for more details and discussions about each failure class). The insets around the pie chart provide exemplary response curves of the circuits in each failure scenario. Each inset shows the concentration of the output node C versus time with two values of the growth-feedback strength kg, one below and another above the failure threshold, for the specific failure scenario. In each case, the input is switched from state I1 to state I2 at the time indicated by the red vertical dashed line.

The three types of continuous failures, denoted as types I-III as illustrated in Figs. 2(a-c), are determined according to the specific quantitative criteria in the response curve that the circuits violate. Type-I continuous failures, as shown in Fig.2(a), are associated with the violation of the precision criterion. A circuit is deemed as precise if a change in the input signal (e.g., from I1 to I2) generates two opposite dynamical effects in the circuit that cancel each other out after a transient and return the final output to the original state, i.e. O2 ≈ O1. For example, in some networks [e.g., the network in Fig. 1(a)], an increase in the input signal I will result in an increase in the concentration of gene A and a reduction in the concentration of gene B. As both genes regulate the output gene C with the respective activation links, for proper system parameter values, the two effects will cancel each other out, resulting in O2 ≈ O1. Type-I continuous failures constitute the largest failure category among all possible circuit topologies, suggesting that the exact cancellation is fragile and the loss of precision is the most common dynamical mechanism behind growth-feedback-induced failures.

Three classes of growth-induced failures

All the failures we observed can be categorized into the following three general classes, applicable to both the three-gene and four-gene circuits we tested:

Continuous Deformation of the Response Curve

Typically, we require a specific range of response curve shapes for a gene circuit, such as a peak in the output with a minimum height or duration. In a failure caused by continuous deformation, the growth feedback prompts a gradual change that crosses the boundary of these criteria for response curve shapes.

Growth-Induced or Growth-Strengthened Oscillations

Growth feedback can induce oscillations in a circuit through various types of bifurcations or amplify existing oscillatory behavior with longer relaxation times or larger amplitudes. A circuit experiencing growthinduced or growth-strengthened oscillations cannot reach a relatively steady state (an equilibrium or relatively small oscillations) within a finite time or reasonable relaxation period.

Growth-Induced Switching Among Coexisting Attractors

When coexisting attractors are present in the circuit dynamics, such as bistability or multistability, the circuit typically only functions with one of the attractors. In other words, the circuit is functional locally in its phase space rather than globally. Strengthened growth can push the system across the boundary of different attracting basins in the circuit phase space, causing the circuit to lose its desired functionality by switching from a functional basin to a malfunctioning basin.

Our simulations reveal that an exact cancellation between the two opposite sources at kg = 0 prevents an exact cancellation at any other values of kg. That is, the set of circuit parameter values leading to perfect precision, in general, depends on the value of kg (see Appendix 5 for more details). The implication is that, for fixed circuit parameter values, achieving high precision under growth feedback (kg > 0) is difficult if the circuit is precise in the absence of growth feedback (kg = 0).

Type-II continuous failures are characterized by a continuous change in the peak of the response curve, denoted as Opeak, as kg increases, eventually falling below a threshold, as shown in Fig. 2(b). This type of failure can make it challenging for downstream circuits to detect the peak signal, hindering information transmission in the larger system. Type-II failures are the second most common type of failure observed in our simulations. The occurrence of a high peak in the response curve requires a significant transient deviation from the final equilibrium point. In the presence of growth feedback, the transient behavior changes, which can further alter the peak height Opeak.

Type-III and type-IV failures arise due to growth-feedback-induced oscillations, while type-V and type-VI failures are caused by bistability or multistability. To provide a more detailed understanding of these different failure scenarios, we discuss the two mechanisms, respectively, in the sections of Growth-feedback induced oscillations and Bistability or multistability below.

Growth-feedback induced oscillations

As demonstrated by the light green and yellow slices of the pie chart in Fig. 2, a considerable portion (17%) of the circuit failures are caused by growth-feedback-induced oscillations. Growth feedback perturbations can easily change the system from the adaptive domain to the oscillation domain in these cases. Our program classifies oscillation-mediated failures into two categories: continuous (type-III) and discontinuous failures (type-IV). Type-III failures are the results of either (i) a gradual increase in the oscillation amplitude, or (ii) a gradual increase in the transient lifetime of damped oscillations. In the first case, an isolated circuit has already exhibited oscillations with small amplitudes in its gene concentrations with relatively weak growth feedback. As the feedback is strengthened with a larger value of kg, the oscillations are intensified with a larger amplitude, leading to a circuit failure. In the second case, there is damped oscillation for small kg with a relatively short transient time before approaching an equilibrium. After strengthening the growth feedback, the damping weakens and the oscillation’s amplitude cannot be reduced to the threshold within the time limit, as exemplified in Fig. 2(c).

The second category of growth-feedback-induced oscillation is type-IV, where oscillations emerge suddenly as the growth-feedback strength increases through a critical point. The sudden emergence of oscillations can be caused by a bifurcation or a transition into a basin of a limit-cycle attractor. A random sampling of the type-IV failure cases reveals that most of them are caused by either a saddle-node bifurcation of cycles (Strogatz, 2018) or an infinite-period bifurcation (Strogatz, 2018). In the former case, a pair of stable and unstable limit cycles suddenly emerge together. In the latter case, when observed from the opposite direction (i.e., with a decreasing kg crossing the threshold), the oscillation in the system spends a longer and longer time around a node on the limit cycle. This node finally becomes a stable fixed point at the bifurcation point, and the oscillation period approaches infinity. One example of type-IV oscillation-mediated failures caused by an infinite-period bifurcation is shown in Fig. 2(d). In our simulations, most of the cases where there are saddle-node bifurcations of cycles are categorized as type-III failures because, prior to the bifurcation point, the system can be oscillating near the ghost cycle (Strogatz, 2018) for a long time exceeding the criterion for relaxation time, though that ghost cycle is not an attractor but only a transient in the system. More details on these two types of bifurcations in our simulated circuits can be found in Appendix 7.

Our results indicate that for various circuit topologies, the dynamic mechanisms leading to failures can differ, resulting in significantly different distributions of failure types among different networks. For instance, the fractions of failures caused by growth-induced oscillations can vary dramatically among all the topologies, as demonstrated in Fig. 3(a), where each data point represents a specific network topology. The fraction of failures caused by growth-induced oscillations can range from approximately zero to about 80%! A particular example of two different networks is presented in Figs. 3(b1) and 3(b2), both of which share the same minimal topology required for adaptation (Shi et al., 2017) - the circuit’s core function. Despite differing by only one link, the proportions of failures with unique mechanisms are quite distinct, as illustrated in Figs. 3(c1) and 3(c2). Notably, for the network in Fig. 3(b1), almost half of the failures result from oscillations, while hardly any oscillation-mediated failures occur for the network in Fig. 3(b2). The explanation is that, although the difference lies in only a single link, this link determines whether an oscillation-correlated motif exists within the network. Previously, three classes of motifs capable of supporting persistent oscillations were discussed (Novák and Tyson, 2008), including the “delayed negative-feedback loop” featured in Fig. 3(b1).

Frequencies of growth-feedback induced oscillation failures for different network topologies.

(a) Significant variations in the proportion of trials resulting in circuit failures due to growth-feedback-induced oscillations (types III and IV failures) across distinct network topologies. Some topologies exhibit virtually no oscillation-related malfunctions, while others experience about 80% of failures caused by growth-induced oscillations. Network topologies containing any oscillation-supporting motifs (discussed in the main text) are represented by red triangles, while the rest are shown as blue circles. The majority of red data points have higher fractions of oscillation-related failures compared to the blue ones, mainly due to the presence of oscillation-supporting motifs. To reduce fluctuations in the results, only circuit topologies with over 200 failed trials are included. (b1, b2) A pair of network topologies that differ by only one link (from node C to node B). (c1, c2) The distinct topologies in (b1, b2) leading to different distributions of failure mechanisms. The topology in (b1) primarily experiences growth-induced oscillation as the major failure mechanism, while the one in (b2) has barely any trials with growth-feedback induced oscillations.

Generally, the circuit dynamics depend sensitively on the structure, but oscillations specifically require a negative feedback loop with time delay (Novák and Tyson, 2008). Since there are no explicit time-delayed terms in the dynamical equations in our model, one of the two types of motifs - an intermediate node in the path of the negative feedback loop or an additional positive feedback loop - is necessary to induce time delay (Novák and Tyson, 2008). For network topologies with a high ratio of functional failures caused by oscillations, both motifs are observed, especially the former type. For the network in Fig. 3(b1), the three links: A → C, C ⊣ B, and B → A, together constitute a negative feedback loop, making the circuit more susceptible to oscillatory behaviors. For the circuit in Fig. 3(b2), no such negative feedback loop exists. Figure 3(a) summarizes the total number of failed trials and the ratio of oscillation-induced failures for each network topology. The network topologies that contain one of the motifs for oscillation as discussed in Novák and Tyson (2008) are marked in red, while the networks that do not consist of any of them are marked in blue. Note that all the networks with relatively high ratios of oscillation-induced failures (e.g., ratio > 0.2) consist of oscillation-correlated motifs. Details about these oscillation-correlated motifs are discussed in Appendix 6.

We conclude that, for a network with an oscillation-correlated motif, even if it is functional at some parameter values, the potential of oscillatory behaviors can be triggered by growth feedback. As a result, networks without these motifs can be safer choices to avoid too many failures cases due to oscillations. Note that this relationship is not deterministic. As shown in Fig. 3, even the networks represented by blue dots that have no oscillation-correlated motifs can still have oscillation-induced failures (with small ratios). The complexity of the scenario makes it challenging to find general and relatively simple rules that connect circuit topology to the circuit’s robustness.

Bistability or multistability

In this section, we describe the dynamical mechanisms behind type-V and type-VI failures, which in total take up about 14% of all the circuit failures. These failures are abrupt, meaning that the response curve undergoes an abrupt change at a certain critical value of kg from a desirable curve of adaptation. Type-V and type-VI failures correspond to an abrupt change in O1 and O2, respectively. Both types of failures are closely related to bistability or multistability.

Bistability and multistability are common phenomena in nonlinear systems. Bistability refers to the situations where two stable attractors coexist in the phase space simultaneously. Multistability describes a similar coexisting phenomenon of attractors, but with more than two attractors. With bistability or multistability in the target dynamical system, the system trajectory may end up in any one of these stable attractors, depending on the initial state of the system evolution. The entire phase space can thus be separated into two or more basins of attraction. Each basin of attraction corresponds to an attractor and consists of all the initial states that eventually lead the system to the attractor. The boundary boundaries separate two different basins of attraction. A close pair of initial states but at different sides of a basin boundary lead the system to different attractors.

In our simulations, we observe bistability in most of the circuit topologies (377 out of 425 circuit topologies). While multistability has also been observed, it is relatively rare. Considering that both bistability and multistability impact the circuits similarly, our subsequent discussions will primarily focus on bistability, which can be straightforwardly extended to multistability. It is highly unusual for both attracting basins to exhibit the desired functional behavior simultaneously. This is because they are located in different regions of the system phase space, and accommodating both would impose overly stringent constraints on the circuit. After all, having one basin functional is already rare enough with a random sampling of circuit parameters. As a result, functional adaptation is typically found in only one of the basins, with adaptation being lost in the other, and the circuit is functional only locally in its phase space, rather than on a global scale. A drifting system parameter, such as kg, can alter the dynamics of the gene circuit. In a situation with bistability, such a change in the system dynamics can modify the shape and position of the basin of attraction and the basin boundary. Consider an initial state close to a basin boundary. With the deformation caused by a drifting parameter, the boundary may shift across the initial state, leading to a sudden switching of the system’s final attractor. If the basin before the parameter change is functional and the basin after is not, this leads to a growth-feedback-induced failure. The crossing of the basin boundary dictates that the system’s final state will abruptly change from one attractor to another. This type of failure can be classified as a switching type of failure.

An example of bistability-related failures is shown in Fig. 2(e), where in the upper panel, the circuit enters into the functional region after an initial transient. In the lower panel, the circuit enters into another region that does not have adaptability, and the circuit does not respond to the switching of the input signal. Figures 4(a) and 4(b) illustrate how the basin structure of the circuit changes significantly with different values of kg. The functional basin is in yellow and it shrinks greatly with an increasing kg. Note that the phase space is four-dimensional, so only a two-dimensional slice is shown. For a bistability/multistability-induced type-V failure where O1 is switched, the boundary of the functional basin crosses the initial state. For a type-VI failure, the simultaneous movement of both O1 and the basin boundary under input I2 results in O1 crossing the basin boundary of O2 states.

Bistability behind both the type-V and type-VI failures.

(a,b) Basin structure of the circuit shown in Fig. 2(e) (with type-V failure) for input I1 with different levels of growth feedback, for growth feedback strength kg = 0.05 (weak) and kg = 0.97 (relatively strong). The coordinates A0 and B0 are the initial values of nodes A and B, respectively, corresponding to a two-dimensional slice of the entire four-dimensional phase space by fixing C = 0.1 and N = 10−3. The color bar indicates the equilibrium value of node C before the input switch, which is O1(C). There is bistability in both cases, as there are two basins of attraction. The yellow region is the functional basin that has adaptation, while the blue region is a non-functional basin without adaptation. The relative size of the blue non-functional region with larger kg in this case is significantly larger and includes the initial state of the system (A0 = B0 = 0.1), causing a type-V circuit failure. (c) Diagram of O2(C) from the circuit in Fig. 2(f) with a type-VI failure. Prior to a threshold value of kg ≈ 0.012, one higher stable value of O2(C) exists. After this threshold, the state suddenly switches to a lower O2(C) one. Note that this abrupt change is not caused by a bifurcation. Instead, it is caused by O1 continuously changing with respect to kg crossing a basin boundary of O2.

One might expect bifurcations to play an important role in many type-V and type-VI failures. However, in our simulations, failures precisely at the bifurcation point are not observed. This is because the bifurcation points under consideration, such as fold bifurcations, are where one of the attraction basins diminishes to zero. For a failure to occur exactly at the bifurcation point, the initial condition would need to coincide precisely with the infinitesimally small basin just before it vanishes. More realistically, failures almost always largely precede the exact bifurcation point. They happen while the basin is still contracting and the basin boundary crosses the initial condition or O1. An example is shown in Fig. 1(b), where bistability persists, yet the lighter orange basin with a larger O1(C) cannot be reached as the boundary shifts away from the initial condition A0 and B0. As another example, in Fig. 1 (c) from a different circuit, the higher O2(C) state disappears at kg ≈ 0.012 and switches to a lower O2(C), but this point is not a bifurcation. It is the point where the stable O1 continuously crosses the basin boundary of O2.

Circuit robustness and optimal topology

To quantify a circuit’s robustness against growth feedback, we introduce two metrics: Q-value and R-value. We track the number of remaining functional trials for each network for various kg values (starting from kg = 0), denoted as Q(kg). This measure extends the concept of Q-values in Ma et al. (2009) by accommodating non-zero values of kg. To characterize the circuit robustness, we define the survival ratio R(kg) as R(kg = k) = Q(kg = k)/Q(kg = 0). This ratio represents the fraction of random circuit realizations that maintain functionality under growth feedback with a strength of kg.

Note that each Q(kg) or R(kg) is defined for a specific network topology in a suitable parameter space. A high value of R(kg) indicates that a large fraction of the randomly sampled circuit parameters is functional despite cell growth with any strength no larger than kg, indicating that the topology is more robust against growth feedback. Because of the detrimental effects of growth feedback, R(kg) decreases monotonically with respect to kg.

To justify the utility of R(kg), we test the circuit topologies employed in a previous work (Zhang et al., 2020), where two relatively simple network topologies were used for a comparison study in terms of their ability to resist growth feedback and remain functional. Our evaluation of R(kg) for the two topologies has yielded results that are consistent with those in Zhang et al. (2020), as discussed in Appendix 2. To illustrate our results in a concrete way, we set kg = 0.6 and calculate the ratio R(kg = 0.6) for different network topologies.

Our computations have revealed a set of eight circuit topologies with optimal performance as characterized by high values of both Q(kg = 0) and R(kg), as indicated by the set of orange points in Fig. 5(a). The optimal circuits form a family as their topologies exhibit a high level of similarity with one other. In particular, all eight circuits in this family share a common set of links (motif), as shown in Fig. 5(b). The combination of these common links is one of the minimal topologies with perfect adaptation in a three regulatory logic (Shi et al., 2017) and is critical for the circuit to be functionally adaptable. The only difference among the circuits in this family is the links from node C. While an inhibition link from node C can be important to achieving a value of R(kg), as discussed below, the eight optimal circuit topologies do not contain any such inhibition link from node C. (The role of this particular link will be further studied in our analysis of the results in Fig. 6.) This also explains why the family has eight members, as follows. Each link from C has two options: either the link does not appear, or it appears as an activation link. As there are three possible links from C (C to A, C to B, and C to C), there are altogether eight (23) topologies within this optimal family, according to the simulation results in Fig. 5.

A family of circuit topologies with optimal performance.

The circuits both have a large volume of the functional region in the parameter space in the absence of growth feedback as characterized by a large value of Q(kg = 0) and are robust against growth feedback with a high value of R(kg). (a) Values of Q(kg = 0) and R(kg = 0.6) from all the 425 network topologies, where each data point corresponds to a topology. The family of optimal topologies is represented by the orange data points, including eight network topologies. (b) The set of links (motif) shared by this family of circuits. The combination of these links is also one of the minimal topologies with perfect adaptation in three regulatory logic (Shi et al., 2017).

Strong correlation between circuit robustness against growth feedback and circuit topology.

There are three groups of circuits, each displaying strong topological similarities within, exhibit distinct levels of robustness against growth feedback as measured by the characterizing quantity R. (a) Robustness measure R(kg = 0.6) versus Q(kg = 0) for all 425 network topologies. Circuits are color/shape-coded into three groups (green triangles, blue circles, and red diamonds) based on the rules defined in the text. The three groups of topologies display distinct levels of R(kg = 0.6) values, signifying a strong correlation between circuit robustness and topology. Only circuits with Q(kg = 0) > 300 are shown to reduce fluctuations arising from random parameter sampling. What is demonstrated is the case of an intermediate level of growth feedback with kg = 0.6 (a different value of kg has no significant effect on the results - see Fig. 7). The topologies associated with the green triangles have a high level of robustness, which can be regarded as an optimal group and is more prevalent than the optimal group identified in Fig. 5. (b) Histogram of R(kg = 0.6) the same color legends as in (a). Three distinct peaks emerge, each associated with a group of circuit topologies. (c) The shared network motif among all networks in the green group, which is highly correlated with the optimal minimal network shown in Fig. 5(b), but without the link B → B, which is necessary for the NFBL family of networks to have adaptation (Shi et al., 2017). (d) Effects of burden b for the three groups of networks, where the abscissa is the effective term of burden in the formula of growth rate Eq. (12). The circuits in the red group have larger values of 1/(1 + ⟨b⟩), suggesting that a heavier burden yields a stronger effect of the growth feedback for the red group. (e) A multilayer perceptron (MLP) for identifying the crucial connections that determine the robustness of the circuits. The circuit topology serves as the input, where 1, 0, and -1 represent activation, null, and inhibition links, respectively. The output is a predicted robustness measure, denoted as . To encourage the neural network to select as few links as possible for predicting , a l-1 regularization term, β ∥Win∥, is incorporated into the loss function alongside the fidelity error . As a result, the feed-forward process eliminates information about the links that have little impact on circuit robustness since the corresponding Win entries automatically optimize to values close to zero. (f) Results from an ensemble of 50 MLPs, each trained with distinct initial values. Shown is the average importance of each of the nine links, which is determined by the weights in Win - see Appendix 8. The top four links with the highest importance correspond to the four links used to classify the three peaks in panel (b).

How can we quickly determine if a three-gene regulatory network with a given topology can be robust against growth feedback? Is there any structural feature of the circuit that can be used to estimate if a high value of R(kg) can be achieved? To gain insights, we observe that the histogram in Fig. 6(b) has three peaks about low, moderate, and relatively high values of R, respectively. Computations reveal certain “shared topological similarity” (or motif) within each peak. Thus, each peak corresponds to a group of similar network topologies that simultaneously have a similar level of R. This observation suggests a correlation between the network topology and robustness against growth feedback. For convenience, we refer to these three groups of networks by the colors presented in Fig. 6. For instance, the group with the highest R (the green triangles in Fig. 6(a)(d)) is called the green group, and the group with the lowest R (the red diamonds in Fig. 6(a)(d)) is the red group.

To better distinguish the three groups, we introduce two binary variables, B1 and B2. For each network, B1 = 1 if the network contains the motif in Fig. 6(c), and B1 = 0 otherwise. Then, for each network, an additional binary variable is set to be B2 = 1 if there is an inhibition link from the output node C, and B2 = 0 otherwise. We find that a linear combination of the two binary variables, BS = B1 − B2, can characterize the circuit topology and robustness against growth feedback. In particular, the three possible cases BS = 0, 1, or -1 correspond to the three peaks in Fig. 6(b). This result suggests that the motif shown in Fig. 6(c) is beneficial for robustness, while an inhibition link from the output node C is detrimental. It is the balancing act of these two factors that determines the overall circuit robustness.

The discovery of this three-peak structure and the corresponding topological similarity within each peak is facilitated with the use of machine learning. In particular, we consider a simple type of artificial neural network called multilayer perceptron (MLP), where we train it to predict the R value from the input of the network topology through a small hidden layer with only two nodes, as demonstrated in Fig. 6(e). This bottleneck structure in the hidden layer plus the l-1 regularization imposed on the input matrix Win forces the MLP to extract low-dimensional features from the input topology to estimate R. In our tests, the MLP designed this way automatically assigns different levels of weights to the input information of different links. Over an ensemble of 50 MLPs trained with different random initial values, the ranking of average importance is shown in Fig. 6 (f). The top four links are the four links used to categorize the three peaks.

The results in Fig. 6 is for kg = 0.6. However, we find that different values of kg lead to essentially the same ranking of R(kg) among the circuit topologies, as illustrated in Fig. 7.

Robustness of the circuit division into three groups subject to different levels of growth feedback.

From the top to the bottom, the four rows are for kg = 0.2, kg = 0.4, kg = 0.6, and kg = 0.8, respectively. The legends are the same as in Figs. 6(a) and 6(b). For different levels of growth feedback, the distribution of the robustness measure exhibits three distinct peaks that occur at approximately the same locations on the R axis. The implication is that the division of the circuit topologies into three groups in terms of the robustness measure can be revealed by examining the circuit functions at a single value of the growth feedback strength.

Three remarks on our categorizing rules based on the two extracted featured motifs are in order.

First, the shared motif for the green group is strikingly similar to the optimal minimal networ in Fig.5(b) (the orange group). The sole distinction lies in the self-activation link of node B. This specific link plays a crucial role. Every network in the NFBL family depends on this link to achieve adaptation (Shi et al., 2017). However, for circuits within the IFFL family, this link is not a necessity for adaptivity. Missing this link makes the motif in Fig. 6(c) no longer a minimal network for adaptation, and a circuit containing this motif may either belong to the NFBL or the IFFL family. We have thus identified two optimal groups: the green group with optimal robustness R and the orange group with both the optimal robustness R and the largest functional volume Q(kg = 0) in the absence of growth feedback. The orange group is a subset of the green group, with an additional requirement for Q(kg = 0).

Second, the shared motif for the red group is also exactly the group of all circuits containing an inhibition link from node C to node B, denoted as C ⊣ B. These two different definitions are in fact equivalent: all networks with Q(kg = 0) > 300 that contain C ⊣ A or C ⊣ C also contain the motif in Fig. 6(c), yielding BS = 0, and belong to the blue group.

Third, the three circuit groups in Fig. 6 are not correlated with the categories used in previous research on circuit functionalities without growth feedback (Ma et al., 2009; Shi et al., 2017). These studies classified adaptive networks into NFBL and IFFL families. Each family contains a few minimal topologies with or without some additional other motifs, and the two families have distinct minimal functional topologies. The minimal topology acts as the backbone for supporting circuit functionality. We find that, when growth feedback is present, the prior classification scheme and the underlying minimal topologies become less relevant. Circuits belonging to the NFBL family are spread across all three levels of R(kg) in Fig. 6(b), as are the circuits from the IFFL family. A robust circuit can be part of either family, just as a fragile circuit can belong to both. We give that: (i) the topological motifs determining circuit functionality robustness and (ii) the motifs deciding whether a circuit belongs to the NFBL or IFFL family are independent. To quantify this irrelevance, we calculate the point biserial correlation between R(kg = 0.6) and a binary variable determining the family to which the circuit belongs. The resulting correlation is merely 0.1, suggesting hardly any correla-tions. A further illustration and quantification of this irrelevance can be found in Appendix 4.

What are the reason and mechanism behind the phenomenological set of circuit categories? Especially, it is desired to understand why the shared motif for the green group is beneficial for circuit robustness, and why the shared motif for the red group is harmful for robustness. It is challenging to find straightforward explanations given the complexity of the problem (see Discussion section). Certain insights are as follows. We find that the average node concentrations at the equilibrium for the network topologies in the red group are consistently smaller than those in the blue group. This difference is reflected in the value of burdens b. In particular, according to Eq. (12), the cell growth rate is proportional to the term 1/(1 + b) under the same level of growth feedback. Figure 6(d) shows the average burden ⟨b⟩ for each network topology, demonstrating that the values of the term 1/(1 + ⟨b⟩) for the circuits in the red group are larger than the values in the blue group. As a result, for the same value of kg, the growth feedback effectively received by the circuits in the red group is stronger than that of the blue group circuits. Further support is provided by the results from the limit J → ∞ (Appendix 3). In this limit, the burden b does not affect the strength of the growth feedback. As a result, the R values of the red group significantly overlap with those of the blue group, suggesting that the distinctively low values of R for the red group be a result of the burden with finite J. We also find that the existence of the shared motif for the red group has a stronger correlation to the motif necessary for growth-feedback induced oscillations. All circuits with oscillation type of failures taking up more than 20% of failures belong to the red group. This correlation can result in further fragility of the red group circuits.

Scaling law quantifying the effect of growth feedback on gene circuits

A comprehensive way to understand the effects of growth feedback on gene circuits is through scaling laws, an approach commonly employed in statistical and nonlinear physics. Does a scaling law exist that characterizes quantitatively how growth feedback affects the circuit functioning? Through a systematic computational analysis of the circuit robustness, we have uncovered a scaling law that governs how the robustness measure R(kg) deteriorates as growth feedback is strengthened, as shown in Fig. 8, where the blue curve is the result averaging over all the 425 network topologies. The three other curves represent circuit topologies that have a relatively high, moderate, and low value R among the 425 topologies tested, to demonstrate that this scaling behavior is generic. (These three topologies are the highest Q(kg = 0) topology in each of the three groups shown in Fig. 6 and Fig. 7.) As growth feedback is strengthened, the number of circuit topologies that can maintain functioning decreases (or, equivalently, the number of failed circuits increases). The decreasing behavior of R(kg) with kg tends to be slower than exponential [e.g., exp(−βkg) with β > 0 being a constant].

Scaling law governing the circuit robustness measure R(kg).

The blue curve is the average result of all the 425 network topologies. The other three curves represent circuits with different robustness levels: high (Circuit No. 98), moderate (Circuit No. 3), and low (Circuit No. 28) values of R, to demonstrate that this scaling behavior is generic. Each of these three circuit topologies is selected from one of the three groups illustrated in Fig. 6 and Fig. 7, and they have the highest Q(kg = 0) value within their respective groups.

A general theoretical argument for the scaling law is unavailable. However, if we simplify the system by setting the parameter J in Eq. (12) to be large so the burden b(t) is much smaller than one, we are able to argue that the scaling law is approximately given by where β > 0 and 0 < λ < 1 are two specific constants that depend on the network topology, and the typical value of λ is about 0.6. The exponential scaling is assumed, given its memorylessness. That is, there is no special zero point of kg, for the reason that a certain level of kg is mathematically equivalent to a larger dx, as discussed below.

The quantity R(kg) is a simple and straightforward measure characterizing the detrimental effect of growth feedback on gene circuits. We carry out a semi-quantitative analysis of this quantity and the effect of kg on it. The circuit dynamical equations Eqs. (9-12) can be simplified by substituting Eq. (14) into them to cancel the dN/dt terms, leading to Compared with the equations without growth feedback Eqs. (6-8), we see that introducing growth feedback is equivalent to adding a variable kg/(1 + b) to the degradation terms for each node. Intuitively, the value Q of a network topology measures the volume of the functional region ℳ in the parameter space, which is also a function of kg. We thus have that R(kg = k) is the volume of the intersection between ℳ (kg = k) and ℳ (kg = 0) divided by the volume of ℳ (kg = 0): where V (ℳ) is the volume of ℳ.

The picture can be further simplified if we assume the burden b is approximately a constant within a range of kg. Since growth feedback contributes to an additional term in degradation dx, strengthening the feedback is equivalent to increasing all three quantities dx together. Consequently, as kg increases, the high dimensional region M does not deform, but simply translates in the negative direction in all dimensions of degradation dx in the parameter space. That is, as growth feedback becomes stronger, it becomes more difficult for the circuit to maintain it functioning.

Discussion

When a synthetic gene circuit is introduced into a host cell, an inherent coupling arises wherein the gene circuit affects cell growth and cell growth in turn alters the circuit gene expression (Klumpp et al., 2009; Klumpp and Hwa, 2014). Due to the fundamental nonlinearity in the gene network and in the cell growth dynamics, the interaction is generally quite complicated. To understand this interaction so as to identify the circuit topologies that can withstand the interaction and maintaining the intended circuit functions is one of the most challenging problems in synthetic biology.

Previous studies showed that growth-mediated feedback can endow synthetic gene circuits with various emergent properties. In general, growth feedback tends to negatively impact the intended function the circuit is designed for. There was preliminary evidence that the effects of growth feedback depend strongly on the circuit topology (Zhang et al., 2020). For a particular circuit function, while the vast majority of the topologies would fall under growth feedback, a handful still exists that is adaptable to maintain its designed functions. Identifying the “optimal” topologies that are most robust against growth feedback is fundamental to constructing synthetic gene circuits that can survive, adapt, and function as designed in the fluctuating growth environment of the host cells.

The main contribution of this paper is a systematic computational study of three-gene circuits with adaptation to uncover and understand the detrimental effects of growth feedback on gene circuits and to identify optimal groups of topologies. Without growth feedback, there are 425 possible topologies with functional adaptation. A vast majority of these circuit topologies fail in their functions under growth feedback, and our computations have revealed, for the first time, six distinct main failure categories covering more than 99% of the cases. From a dynamical point of view, there are three mechanisms by which growth feedback can deprive the circuit of its ability to adapt:

continuous deformation of the response curve, (ii) strengthened or induced oscillations, and (iii) sudden switching to coexisting attractors (also summarized in Box 1). By introducing a robustness measure to quantify circuit adaptation in the presence of growth feedback, we uncover a general scaling law characterizing the detrimental effect of growth feedback on the circuit functioning in a quantitative manner. We identify an optimal group of circuits with high robustness and key subsets of links associated with this group that play a critical role in sustaining circuit function in host cells. Taken together, to design a functional gene circuit, growth feedback must be taken into account, as the same circuit designed with perfect functions without the feedback can behave quite differently when the feedback is present. Our study has provided unprecedentedly quantitative insights into the interplay between gene circuit topology and growth feedback, unlocking the dynamical mechanism of growth-induced failures and providing guidance to better design practically applicable synthetic gene circuits.

A unique finding is that growth feedback can induce or strengthen oscillations in gene circuits designed for adaptation. Such oscillations can often destroy the circuit functionality. In a recent experimental study, a similar phenomenon was observed in gene circuits designed for selfactivation (Melendez-Alvarez et al., 2021). These results suggest that growth-feedback induced oscillation may be a general dynamical mechanism that can negatively affect the robustness of gene circuits. In addition, our study has shown that growth feedback has a highly sensitive dependence on the circuit topology: even a small structural differences between two circuits designed for the same function can result in drastically different outcomes under growth feedback. For example, Fig. 3 demonstrates that a link critical for an oscillation-supporting motif can significantly affect the robustness of the circuit against growth feedback. It can thus be quite useful to identify failure-related motifs so that they can be avoided when designing a gene circuit.

From a broad point of view, our study has yielded basic insights into the fundamental topologyfunction relationships in gene circuits. Examples include how circuit topology affects circuit robustness against growth feedback and whether a circuit topology contains motifs supporting a specific type of growth-induced failure, such as oscillation-related malfunctions. However, searching for and understanding the interplay between circuit topology and dynamical behaviors remain to be a challenge, for the following five reasons.

First, the two relevant questions are whether a circuit topology supports adaptation and whether the circuit is robust against growth feedback or is susceptible to a specific type of growth-induced failures. While our study focused on the latter, the former is important. Addressing both questions to identify and analyze all possible scenarios is infeasible at the present, due to the complex parameter space of the circuits. To make our study feasible, we focused on the cases where the circuit satisfies all the requirements for adaptation in the absence of growth feedback. These cases may occupy a small region in the entire parameter space of the circuit. For each circuit topology, the uncovered function failures due to growth feedback are thus limited to relatively small parameter regions. Second, most network topologies studied have dense connections among the three nodes (only about 20% of the networks have fewer than six connections). As a result, different motifs can overlap with each other, blocking or enhancing the function of each other. The dense connections thus pose a difficulty in identifying the motifs accurately. Third, for a particular class of failures, competition among different failure types may arise. For instance, a circuit with oscillation-supporting motifs may not have a high fraction of oscillation-induced failures because it also contains the motif for bistability, leading to a large fraction of failures due to the bistability-induced malfunctions. Fourth, due to the necessity to set a threshold in the relaxation time, transient behaviors can arise. In many failure cases caused by oscillations, the oscillatory behavior is not stable and the circuit will eventually approach a fixed point. However, time scales should be taken into account. The transient behaviors can make the network topologies without the necessary motif for sustained oscillations exhibit oscillation-induced failures. Fifth, growth feedback acts as additional feedback loops within the circuit, potentially complicating the circuit dynamics and adding more links to the circuit topology. These extra links in the integrated topology might give rise to an oscillation-related motif. However, our simulations have shown that the impact of this additional oscillation motif, introduced by growth feedback, tends to be weak (Appendix 3).

Although the primary focus of this paper is on how growth feedback can undermine an originally adaptive circuit and how to design circuits that are robust against such feedback, our simulated dataset reveals instances where growth feedback can benefit the circuit within certain ranges. Specifically, we identified 2,092 circuits across 306 different topologies where adaption, lost at an intermediate level of growth feedback, is restored at higher levels. This is 1.4% of all circuits tested. We anticipate that additional circuits exhibiting this loss-and-recovery behavior exist, as our sampling of six discrete levels of kg (0,0.2,0.4,0.6,0.8,1.0) might have overlooked numerous cases. This result again suggests the possible advantages of growth feedback in gene circuits (Tan et al., 2009; Nevozhay et al., 2012; Deris et al., 2013; Feng et al., 2014). A comprehensive study into how growth feedback can endow or enhance adaption in circuits would require entirely different approaches for sampling circuit parameters and selecting candidate network topologies, demanding significantly high computational costs. Given that this topic extends beyond the scope of the current paper, we leave this matter to future research.

Our study focuses on scenarios where random noises are ignored. Realistically, gene circuits are subjected to diverse types of noise, which can complicate their predictable behavior and design. These noises can originate externally from a noisy input signal I, or intrinsically, directly affecting the circuit components. Further, these noises can be classified based on various mechanisms that cause them Colin et al. (2017); Sartori and Tu (2011). And with different mechanisms, each type of noise can be characterized by different attributes such as frequency, amplitude, and noise color. These variances can lead to different impacts on the circuits, potentially necessitating unique mechanisms or designs for the attenuation of each category Sartori and Tu (2011); Qiao et al. (2019). Given the extensive complexity and the need for thorough investigation, these noise-related challenges are beyond the scope of this paper and require a series of future studies.

In our paper, we consider dilution due to cell growth as the dominant factor of growth feedback. Here we compared the adaptive circuits under no-growth conditions and their ability to maintain their adaptive behaviors after dilution into a fresh medium, which mediated a significant dilution to the circuits. This is based on our previous work Zhang et al. (2020). However, growth feedback is inherently complex Klumpp et al. (2009). For instances, an increased growth rate can change protein synthesis rate Hintsche and Klumpp (2013); Scott and Hwa (2023), and cell growth rates can affect the distribution of protein expression in cell populations Gouda et al. (2019). In our paper, we concentrate on a simplified model with dilution, which we consider to have captured the dominant factor. The dynamic roles of the dilution and growth-affected production rate should be analogous, given that they both act as inhibitory factors arising from cell growth. Incorporating the impact of growth rate on protein synthesis into our model would offer a more comprehensive analysis, a task beyond the scope of this paper but presenting an intriguing opportunity for future research to address the complexities of growth feedback.

In Appendix 1, we extend our analysis to four-gene circuits with over two thousand functional failure trials. A remarkable finding is that the failure scenarios for these four-gene circuits are the same as the categories for three-gene circuits (summarized in Box 1), indicating that the growth-feedback induced failure mechanisms identified in our work are general. The focus of our study on small gene circuits is driven by their current relevance in synthetic biology. The primary reason is that even modest-sized circuits, when introduced to a host, can provoke unintended and often uncontrollable outcomes due to competition and interactions in the form of growth feedback. Additionally, resource competition within the host cell can arise, where circuit genes compete for limited resources, adversely affecting the dynamics of the circuit. Consequently, larger gene circuits are generally not favored in synthetic biology due to these complexities. Indeed, the state-of-theart synthetic gene circuits typically involve only three to four genes, a realm where the implications of growth feedback have been insufficiently understood until now. Our work aims to bridge this knowledge gap.

It is possible that, in the future, synthetic biology may use larger and more complex circuits. To uncover and understand the failure mechanisms as well as to identify circuits that are resilient to growth feedback, machine learning can be used. For example, recurrent neural networks have recently been used to identify circuit topologies appropriate for a specified desired function (Shen et al., 2021), and reinforcement learning tackle the combinatorial optimization problem (Bello et al., 2016; Mazyavkina et al., 2021) of pinpointing the optimal circuit topologies. Furthermore, automated differentiation (Hiscock, 2019; Kong, 2022) can be exploited to locate optimal network parameters, which can be efficient for larger circuits with a high-dimensional parameter space. In spite of these works, to study the effects of growth feedback and resource competition among numerous genes in larger circuits remains to be a formidable challenge. Our work providing a comprehensive picture of the failure mechanisms induced by growth feedback represents a step forward in this field.

Models and Methods

Model description

We restrict our study to the class of transcriptional regulatory networks (TRNs) with the AND logic. In order not to overwhelm readers with too many terms and parameters, we first describe a partial model (an isolated circuit without growth feedback) before introducing the complete model that we study in this work. For an isolated circuit without any growth feedback, and with the topology specified inside the red dashed box in Fig. 1(a), the dynamical equations are where the dynamical variables A, B, and C are the concentrations of each protein (node). The notations are as follows. Let x and y be two arbitrary nodes. The quantity υx is the maximal production rate of gene x, dx is the degradation rate of gene x, dx/dt is its time derivative of the concentration, nxy and Kxy are the coefficients in the Hill function for a transcriptional regulation from gene x to gene y.

When growth-mediated feedback is present, the dynamical equations of the three-node circuits are modified to where the additional dynamical variable N denotes the density of the host cells, kg is a parameter controlling the maximal growth rate of the host cells, J is a parameter reflecting how this threenode gene circuit contributes to the burden. Equations. (9) to (13) are the dynamical equations we actually use for simulating the circuit dynamics.

The growth of N is under the regulatory action of two sources: by itself following the logistic equation with the environmental capacity N0 and by the burden b that represents the competence from the metabolism of the gene circuit. To make the computations feasible, we focus our analysis on the exponential growth phase so that N0 ≫ N. The equation governing the growth of the cell numbers, Eq. (12), can then be rewritten as where the dilution rate dN/dt is regulated only by the burden b(t) of the gene circuit. While cell growth is inhibited by the metabolism of the gene circuit, the circuit is also regulated by the growth of N that dilutes the concentration of circuit nodes with increasing cell volume. This dilution is reflected by the additional terms −(x/N)(dN/dt) in Eqs. (9-11).

It is useful to clarify the meaning of the degradation parameter dx and its relationship to growth feedback. While degradation and growth feedback terms have the same sign in the regulatory equations, dx may include a constant dilution. We assume that dx represents the sum of all the degradation effects in cells that are distinct from growth feedback. For instance, degradation tags, especially in the ssrA tagging systems (Gottesman et al., 1998), are often used in synthetic gene circuits to increase the degradation rate and thus increase the time scale of the whole system (Elowitz and Leibler, 2000; Fung et al., 2005; Stricker et al., 2008; O’Brien et al., 2012).

Numerical simulations of circuit dynamics

We use the fourth-order Runge-Kutta method to numerically integrate the dynamical equations of the gene circuits, with the integration step as tstep = 0.05. The dynamical equations we use are similar to Eqs. (9-13) but with different topologies. All the initial states of A, B, and C are taken to be 0.1. The input signal is initially I0 = 0.06 and then switched to I1 = 0.6. The simulation codes can be found in our GitHub repository (link provided in Code availability). The simulation results can be found in our OSF repository (link provided in Data availability).

To achieve the desired adaptation, the circuit’s output should reach a steady state before and after the input signal is switched. The values of O1 and O2 can be determined as the output signal associated with the steady states. However, realistically, it is not necessary for the circuit to reach an exact equilibrium. From a computational perspective, the system’s state is almost always asymptotically approaching that exact equilibrium but not actually reaching it. As we are simulating the dynamical process of the circuit, setting a condition such as dA/dt = dB/dt = dC/dt = 0 to determine O1 or O2 is not meaningful. From a more biological perspective, relatively small drifting or oscillations in the circuit should not harm the circuit functionality and are acceptable. Therefore, we define a “relatively steady state” where, within a time block of tblock = 200, the standard deviation of the time series of each node x(t) satisfies: std(x) < 1 × 10−4 and std(x)/mean(x) < 0.05. To further guarantee that the circuit is actually in the “relatively steady state,” two successive time blocks satisfying the standard deviation requirements are needed. The quantities O1 or O2 are then defined as the respective mean values of the output signal in that last time block tblock.

Numerical criteria for functional adaptation and failure types

We introduce four criteria to determine if a circuit has functional adaptation, and the failure types. The codes implementing these criteria are available in our GitHub repository, with the link provided in Code availability. The failure type results for all circuits tested are available in our OSF repository, with the link provided in Data availability. An additional note is provided in the README file of our GitHub repository for further guidance on generating pie charts similar to Fig. 2 for any network topology or subset of topologies.

Precision

The basic requirement of adaptation is that the output remains the same when is input is switched from one state to another, i.e., O2 should be close to O1 in Fig. 19b). Specifically, we set the precision criterion to be |(O2 − O1)/O1|< 0.1.

Sensitivity

The circuit is also required to respond to the switch of the input signal with a high peak. This ability of the circuit is named sensitivity. We introduce two types of sensitivity: relative and absolute, with the respective criteria Opeak/O1 > 0.5 and Opeak > 0.1. Only the circuits meeting both criteria are regarded as having achieved the required sensitivity.

The need to use the two different criteria simultaneously can be justified, as follows. Given the variety of network topologies and a large number of system parameters, there is a vast diversity in the circuit dynamics and the values of O1. When O1 is small, it is difficult to observe a peak that has even satisfied the relative sensitivity criterion. If the absolute criterion is used alone for a circuit with a large O1 value, the peak may be negligible in comparison with O1, making its observation practically difficult. It is thus necessary to combine the two criteria so that the cases of small and large values of O1 can be dealt with on the same footing.

Oscillations and relaxation time

An ideal gene circuit should be able to respond and adapt within a reasonable time scale. We set an upper bound of evolution time tmax = 4, 000. If the system cannot reach the “relatively steady state” within this time, it is regarded as non-functional. According to our results, a circuit exceeds our relaxation time upper bound almost always due to oscillations.

Continuous or abrupt failures

We categorize failures into continuous failures and abrupt switching, as exemplified in Fig. 2. After determining the circuit’s critical kg value for a circuit, we calculate the changes in O1 and O2 both before and after this critical point to evaluate if the change is continuous or not across the critical value. More specifically, taking O1 as an example, we calculate the O1 values before and after the critical value, O1,before and O1,after. Then, for each gene A, B, and C, we calculate their changes in O1,before and O1,after. Taking gene A as an example, if O1,before(A) > 0.001, then a relative difference|O1,after (A) – O 1,before (A)|/O 1,before (A) > 0.05 is considered as abrupt. If O1,before(A) < 0.001, the relative difference is not reliable as when A approaches zero, it can be observed as A ≈ 10−10 or A ≈ 10−12, depending on the rather arbitrary time window. So in this case, when O1,before(A) < 0.001, we use a criterion based on the absolute difference |O1,after(A) − O1,before(A)|> 0.01. If any of the genes have abrupt switching, then the entire failure is considered as an abrupt switching.

Details of parameter space sampling

A three-node gene circuit subject to growth feedback has a large number of parameters. Let L be the number of links among the three nodes (excluding the input link). The total number of parameters is 2 · 3 + 2(L+ 1) = (2L+ 8). The values of these parameters determine the properties of the regulation links within the circuit and, as a result, the circuit dynamics. The circuit parameters are randomly generated by the Latin hypercube sampling method (Iman et al., 1980) using the function “lhsdesign” in Matlab. The parameters are sampled uniformly either on a logarithmic or a linear scale. The sampling ranges of the parameters are: υx ∈ [10−1, 101] (sampled in logarithmic scale), dx ∈ [10−2, 1] (sampled in logarithmic scale), nxy ∈ [1, 4] (sampled in linear scale), and Kxy ∈ [10−3, 1] (sampled in logarithmic scale).